---
title: "Class Project"
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document: default
---


### **This landing page displays the knitted output of our README.Rmd file. For the code behind the analysis and figures shown below, please consult the README.Rmd file.**

# Repository Guide

This repository contains our group project for Software Tools for Data Analysis. Our project studies how mortgage interest rates and housing prices affect homeownership rates across U.S. states over time. The final audience for this work is Scott Turner, U.S. Secretary of Housing and Urban Development (HUD).

### Branch Structure
- `main`: current working branch for the final project deliverables
- `Checkpoint-1` branch: preserved as a historical archive of our first checkpoint submission
- `Checkpoint-2` branch: preserved as a historical archive of our second checkpoint submission

### Repository Contents
- `README.Rmd`: source file for this repository landing page
- `README.md`: knitted GitHub-facing version of the README
- `Data/`: raw and cleaned data files used in the analysis
- `Report Outline.docx`: working outline for the written report
- `README_files/`: figures generated from the README

### Team Workflow
To keep the repository organized, team members should:
1. Pull the most recent version of the repository before starting work
2. Make changes locally in the appropriate files
3. Knit `README.Rmd` after updating it so that `README.md` stays current
4. Commit changes with clear commit messages
5. Push updates after meaningful progress

### Current Project Responsibilities
- Ryan: Introduction, Data Summary, README/repository instructions
- Levi: Data Analytics
- Liz: Conclusion, Policy Recommendations, repository/class page coordination

This README serves as the main landing page for our project progress, documentation, meeting notes, and selected analysis outputs.


```{r clear_environment, include=FALSE}
remove(list=ls())
```

# Introduction


Homeownership remains one of the most important pathways to wealth-building, stability, and long-term financial security in the United States. However, recent housing market conditions have made homeownership increasingly difficult to attain for many families and individuals. Higher mortgage rates raise the cost of borrowing, while rising housing prices increase the upfront and long-term financial burden of purchasing a home. These pressures are especially important for the U.S. Department of Housing and Urban Development (HUD), which is responsible for supporting fair, affordable, and sustainable access to housing across the country.

This report is written for Scott Turner, U.S. Secretary of Housing and Urban Development. As HUD considers how to address challenges with affordability, expand access to homeownership, and support stable housing markets, policymakers need evidence on which factors are most closely associated with changes in homeownership rates. While mortgage rates and housing prices are central to the current affordability discussion, homeownership is also shaped by broader state-level conditions. Some of these include income, unemployment, rent burden, and housing supply. For that reason, our analysis examines both the direct relationship between borrowing costs, housing prices, and homeownership, as well as the role of additional economic and housing market drivers.

Our research question is: **How do mortgage interest rates, housing prices, and related state-level economic and housing market conditions affect homeownership rates across the United States over time?** To answer this question, we construct a state-year panel dataset covering U.S. states from 2005 to 2024. We exclude 2020 due to comparability issues in one-year American Community Survey data during the COVID-19 period. We combine public data from the U.S. Census Bureau, Federal Reserve Economic Data (FRED), the Federal Housing Finance Agency (FHFA), the Bureau of Labor Statistics (BLS), and the Census Building Permits Survey.

The results of this analysis can help HUD evaluate which policy levers may be most relevant for improving access to homeownership. If homeownership rates are strongly related to mortgage rates, housing prices, rent burden, or housing supply, then federal housing policies can be better targeted toward fixing conditions that most constrain citizens from purchasing homes. These decisions have consequences beyond individual buyers: homeownership patterns affect household wealth accumulation, neighborhood stability, housing market equity, and the broader economic health of communities.


# Data Summary

To study these relationships, we construct a state-year panel dataset covering U.S. states from 2005 to 2024, excluding 2020 because one-year American Community Survey data were not available in a comparable form during the COVID-19 period. Each observation in the dataset represents one state in one year. This structure allows us to compare both differences across states and changes within states over time, making it well suited for analyzing how economic and housing market conditions are associated with homeownership outcomes.

The dataset combines several reputable public sources. Homeownership rates, median income, population, and rent burden are drawn from the U.S. Census Bureau’s American Community Survey. Mortgage rates come from the Federal Reserve Economic Data database, while state-level housing price data come from the Federal Housing Finance Agency House Price Index. We also include unemployment data from the Bureau of Labor Statistics and housing permit data from the Census Building Permits Survey as a proxy for housing supply. After importing, cleaning, and merging these sources, we use the resulting dataset to examine patterns in homeownership, affordability, labor market conditions, and housing supply across states. This combined dataset provides the foundation for both our descriptive analysis and our regression modeling.


```{r load_packages, message=FALSE, warning=FALSE, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))
# Load required packages
packages <- c(
  # data manipulation
  "tidyverse",
  "janitor",
  # data import/extraction
  "readr",
  "readxl",
  "car",
  "tidycensus",
  "fredr",
  # dates and time handling
  "lubridate",
  # time series structure and analysis
  "tsibble",
  "fpp3",
  "patchwork",
  # regression
  "sandwich",
  "lmtest",
  "stargazer",
  "modelsummary"
)

# checks if package available (if not - install packages)
# load packages
check.packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}
check.packages(packages)
```


```{r set_api_keys, include=FALSE}
### **Set API Keys:**
#tidycensus api key: https://api.census.gov/data/key_signup.html
#fredr api key: https://fred.stlouisfed.org/docs/api/api_key.html

census_api_key("894b3464183cec8f6c0fa053eb34d0d878de682e")
fredr_set_key("8351bf4f459d0e8c92d56f50021947d8")
```


```{r import_data, message=FALSE, warning=FALSE, include=FALSE}
#Set years to include the last two decades, excluding 2020 due to the COVID-19 pandemic.
years <- c(2005:2019, 2021:2024)

#Extract home ownership data from the Census
homeownership <- map_dfr(
  years,
  function(yr) {
    get_acs(
      geography = "state",
      variables = c(
        total_occupied = "B25003_001",
        owner_occupied = "B25003_002"
      ),
      year = yr,
      survey = "acs1"
    ) %>%
      select(NAME, variable, estimate) %>%
      pivot_wider(names_from = variable, values_from = estimate) %>%
      mutate(
        year = yr,
        homeownership_rate = 100 * owner_occupied / total_occupied
      ) %>%
      transmute(
        state = NAME,
        year,
        homeownership_rate
      )
  }
)

#Extract income data from the Census
income_list <- map(
  years,
  ~ get_acs(
      geography = "state",
      variables = "B19013_001",
      year = .x,
      survey = "acs1"
    ) %>%
      transmute(
        state = NAME,
        year = .x,
        median_income = estimate
      )
)

#Bind the income data into a DF
income <- bind_rows(income_list)

#Extract population data from the Census
population_list <- map(
  years,
  ~ get_acs(
      geography = "state",
      variables = "B01003_001",
      year = .x,
      survey = "acs1"
    ) %>%
      transmute(
        state = NAME,
        year = .x,
        population = estimate
      )
)

#Bind the population data into a DF
population <- bind_rows(population_list)

# Rent burden from ACS table B25070
# Defined as share of renter households paying 30% or more of income toward gross rent
rent_burden <- map_dfr(
  years,
  function(yr) {
    get_acs(
      geography = "state",
      table = "B25070",
      year = yr,
      survey = "acs1"
    ) %>%
      select(NAME, variable, estimate) %>%
      mutate(year = yr) %>%
      pivot_wider(names_from = variable, values_from = estimate) %>%
      transmute(
        state = NAME,
        year,
        rent_burden = 100 * (
          B25070_007 + B25070_008 + B25070_009 + B25070_010
        ) / B25070_001
      )
  }
)

# Housing supply from Census Building Permits Survey
# Read directly from Census-hosted Excel file
permits_url <- "https://www.census.gov/construction/bps/xls/annualhistorybystate.xls"

tmp_permits <- tempfile(fileext = ".xls")
download.file(permits_url, tmp_permits, mode = "wb")

sheet_names <- excel_sheets(tmp_permits)
sheet_names <- sheet_names[sheet_names != "PR"]

permits_state <- map_dfr(
  sheet_names,
  function(sh) {
    read_excel(
      tmp_permits,
      sheet = sh,
      skip = 5,
      col_names = FALSE
    ) %>%
      select(1:5) %>%
      setNames(c(
        "universe",
        "year",
        "total_buildings",
        "housing_permits",
        "total_valuation"
      )) %>%
      filter(
        !is.na(year),
        str_detect(as.character(year), "^[0-9]{4}$")
      ) %>%
      transmute(
        state = state.name[match(sh, state.abb)],
        year = as.integer(year),
        housing_permits = readr::parse_number(as.character(housing_permits))
      ) %>%
      filter(year %in% years)
  }
)

#Extract mortgage data from FRED
mortgage_raw <- fredr(
  series_id = "MORTGAGE30US",
  observation_start = as.Date("2005-01-01"),
  observation_end = Sys.Date()
)

#Estimate annual mortgage rates
mortgage_annual <- mortgage_raw %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarise(
    mortgage_rate = mean(value, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r import_local_data, message=FALSE, warning=FALSE, include=FALSE}
# Import Local Data
fhfa_raw <- read_csv("Data/hpi_at_state (1).csv", col_names = FALSE) %>%
  setNames(c("state", "year", "quarter", "hpi"))

# Set HPI data by state/year
fhfa_annual <- fhfa_raw %>%
  clean_names() %>%
  filter(year >= 2005) %>%
  group_by(state, year) %>%
  summarise(
    hpi = mean(hpi, na.rm = TRUE),
    .groups = "drop"
  )

hpi_annual <- fhfa_annual %>%
  mutate(state = state.name[match(state, state.abb)])

# Import local unemployment data
bls_data <- read_delim("Data/la.data.2.AllStatesU.txt", delim = "\t") %>%
  clean_names()

bls_series <- read_delim("Data/la.series.txt", delim = "\t") %>%
  clean_names()

bls_area <- read_delim("Data/la.area", delim = "\t") %>%
  clean_names()

# Join the actual data to the series
bls_merged <- bls_data %>%
  left_join(bls_series, by = "series_id")

# Use key to select code
bls_merged <- bls_merged %>%
  left_join(
    bls_area %>% select(area_code, area_text),
    by = "area_code"
  )

# Keep annual-average unemployment rate
state_unemployment <- bls_merged %>%
  filter(period == "M13") %>%          # annual average
  filter(measure_code == "03") %>%     # unemployment rate
  transmute(
    state = area_text,
    year = as.integer(year),
    unemployment_rate = as.numeric(value)
  ) %>%
  filter(state %in% c(state.name, "District of Columbia")) %>%
  arrange(state, year)
```



```{r merge_data, message=FALSE, warning=FALSE, include=FALSE}
#Merge the data sets by year and state
merged <- left_join(
  population,
  income,
  by = c("state", "year")
) %>%
  left_join(
    homeownership,
    by = c("state", "year")
  ) %>%
  left_join(
    hpi_annual,
    by = c("state", "year")
  ) %>%
  left_join(
    state_unemployment,
    by = c("state", "year")
  ) %>%
  left_join(
    rent_burden,
    by = c("state", "year")
  ) %>%
  left_join(
    permits_state,
    by = c("state", "year")
  ) %>%
  left_join(
    mortgage_annual,
    by = "year"
  )
```


```{r permits_per_person, message=FALSE, warning=FALSE, include=FALSE}
#Estimate annual mortgage rates
merged <- merged %>%
  mutate(perm_per_person = housing_permits/population, NA)
```


```{r excludeDCandPR, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Filter out D.C. and PR
merged <- merged %>%
  filter(!state %in% c("Puerto Rico", "District of Columbia"))

#Remove a bad column
merged <- merged[, names(merged) != "NA"]
```


```{r pop_summary_stats, echo=FALSE, warning=FALSE, message=FALSE}
#Make a df version of merged
merged_df <- as.data.frame(merged)

#Rename quantitative variables in merged_df
merged_df <- merged_df %>%
       select('Population' = population,
              "Median Income" = median_income,
              "Homeownership Rate" = homeownership_rate,
              "Mortgage Rate" = mortgage_rate,
              "Housing Price Index" = hpi,
              "Unemployment Rate" = unemployment_rate,
              "Rent Burden" = rent_burden,
              "Housing Permits" = housing_permits,
              "Housing Permits Per Person" = perm_per_person)

#Compute summary statistics for all quantitative variables
datasummary_skim(merged_df[, sapply(merged_df, is.numeric)], histogram=FALSE, title = "Summary Statistics", align="c")

```


[Discussion of summary stats.]


### **Plot Histograms of Key Variables:**
```{r figure_1, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=7.5}
key_vars_labs <- c("homeownership_rate" = "Homeownership Rate", "hpi" = "House Price Index", "median_income" = "Median Income", "population" = "Population", "rent_burden" = "Rent Burden", "unemployment_rate" = "Unemployment Rate", "perm_per_person" = "Housing Permits per Person")

#Create histogram plots of key variables
merged %>%
  select(
    homeownership_rate,
    median_income,
    population,
    hpi,
    unemployment_rate,
    rent_burden,
    perm_per_person
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 2, 
             labeller = as_labeller(key_vars_labs)) +
  labs(
    title = "Histograms of Key Variables",
    x = NULL,
    y = "Count"
  ) +
  theme_minimal()
```

[Discussion of histograms.]

# Data Analytics

### **Key Trends Over Time:**

```{r figure_2, echo=FALSE, message=FALSE}
#| fig-width: 10
#| fig-height: 7
# Average homeownership rate over time
avg_homeownership_ts <- merged %>%
  group_by(year) %>%
  summarise(
    avg_homeownership_rate = mean(homeownership_rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as_tsibble(index = year)

p1 <- autoplot(avg_homeownership_ts, avg_homeownership_rate) +
  labs(
    title = "Average State Homeownership Rate Over Time",
    x = "Year",
    y = "Homeownership Rate (%)"
  ) +
  theme(
    plot.title = element_text(size = 18),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.margin = margin(t = 6, r = 10, b = 6, l = 10)
  )

# Average HPI over time
avg_hpi_ts <- merged %>%
  group_by(year) %>%
  summarise(
    avg_hpi = mean(hpi, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as_tsibble(index = year)

p2 <- autoplot(avg_hpi_ts, avg_hpi) +
  labs(
    title = "Average State House Price Index Over Time",
    x = "Year",
    y = "House Price Index"
  ) +
  theme(
    plot.title = element_text(size = 18),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.margin = margin(t = 6, r = 10, b = 6, l = 10)
  )

# Mortgage rate over time
mortgage_ts <- mortgage_annual %>%
  as_tsibble(index = year)

p3 <- autoplot(mortgage_ts, mortgage_rate) +
  labs(
    title = "Average 30-Year Mortgage Rate Over Time",
    x = "Year",
    y = "Mortgage Rate (%)"
  ) +
  theme(
    plot.title = element_text(size = 18),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.margin = margin(t = 6, r = 10, b = 6, l = 10)
  )

# Combine all three plots
p1 / p2 / p3
```
Figure 2 shows broad time trends in average state homeownership rates, house price index values, and 30-year mortgage rates. Average homeownership declined from the mid-2000s through the mid-2010s, reaching its lowest point around 2014–2015 before gradually recovering in the late 2010s and early 2020s. Over the same period, the average state house price index increased substantially, especially after 2020, suggesting that housing prices rose much faster than homeownership rates recovered.

The mortgage rate trend provides additional context. Mortgage rates generally declined from the mid-2000s through 2021, but then rose sharply in the early 2020s. This creates an important policy concern, as even though homeownership partially recovered after the mid-2010s, households now face a combination of elevated housing prices and higher borrowing costs. Together, these trends suggest that homeownership affordability is shaped by multiple pressures at once rather than by any single variable alone.

### **Correlation Matrix:**
```{r figure_3, echo=FALSE, message=FALSE}
#Create correlation matrix
#Select variables of interest
corr_data <- merged %>%
  select(
    homeownership_rate,
    median_income,
    population,
    hpi,
    unemployment_rate,
    mortgage_rate,
    rent_burden,
    perm_per_person
  )

corr_matrix <- cor(corr_data, use = "complete.obs")

corr_df <- as.data.frame(as.table(corr_matrix))

# labels
var_labels <- c(
  homeownership_rate = "Homeownership Rate",
  median_income      = "Median Income",
  population         = "Population",
  hpi                = "House Price Index",
  unemployment_rate  = "Unemployment Rate",
  mortgage_rate      = "Mortgage Rate",
  rent_burden        = "Rent Burden",
  perm_per_person    = "Housing Permits per Person"
)

# keep only upper triangle, relabel
corr_df <- corr_df %>%
  mutate(
    Var1 = factor(Var1, levels = colnames(corr_matrix)),
    Var2 = factor(Var2, levels = colnames(corr_matrix))
  ) %>%
  filter(as.numeric(Var1) <= as.numeric(Var2)) %>%
  mutate(
    Var1 = factor(var_labels[as.character(Var1)], levels = var_labels),
    Var2 = factor(var_labels[as.character(Var2)], levels = var_labels)
  )

#Plot the correlation matrix
ggplot(corr_df, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", Freq)), size = 3.5) +
  scale_fill_gradient2(
    low = "firebrick",
    mid = "white",
    high = "steelblue",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation Heatmap of Key Variables",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

Figure 3, the correlation heatmap provides an initial look at the relationships among different variables before moving into regression analysis. Homeownership has negative correlations with population, rent burden, house price index, median income, and unemployment, suggesting that states with larger populations, higher housing costs, and greater rental affordability pressure tend to have lower homeownership rates. The negative relationship with rent burden is especially important because it suggests that places where renters face greater financial strain may also be places where households have more difficulty transitioning into ownership.

The heatmap also shows several relationships among the explanatory variables themselves. Median income and house price index have a strong positive correlation, indicating that higher-income states also tend to have more expensive housing markets. Rent burden is positively correlated with population and house price index, which supports the broader affordability story that larger and more expensive states tend to place more pressure on renters. These correlations help motivate the use of multivariable regression, since homeownership is related to several overlapping economic and housing market conditions rather than one factor in isolation.

```{r missing_check, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Check for missing obsevations
merged %>%
  summarise(
    across(
      c(
        homeownership_rate,
        median_income,
        population,
        hpi,
        unemployment_rate,
        mortgage_rate,
        rent_burden,
        perm_per_person
      ),
      ~ sum(is.na(.))
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  )
```


```{r figure_4, echo=FALSE, message=FALSE, }
# Create lines of best fit for key variables

# Identify key variables
kpi_labs <- c(
  "hpi" = "House Price Index",
  "median_income" = "Median Income", 
  "mortgage_rate" = "Mortgage Rate", 
  "perm_per_person" = "Housing Permits per Person", 
  "rent_burden" = "Rent Burden", 
  "unemployment_rate" = "Unemployment Rate"
)

# Plot the lines of best fit
merged %>%
  select(
    homeownership_rate,
    mortgage_rate,
    hpi,
    unemployment_rate,
    median_income,
    rent_burden,
    perm_per_person
  ) %>%
  pivot_longer(
    cols = -homeownership_rate,
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value, y = homeownership_rate)) +
  geom_point(alpha = 0.35, size = 1.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 0.9) +
  facet_wrap(
    ~ variable,
    scales = "free_x",
    ncol = 2,
    labeller = as_labeller(kpi_labs)
  ) +
  labs(
    title = "Homeownership Rate vs. Key Predictors",
    x = NULL,
    y = "Homeownership Rate (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16),
    strip.text = element_text(size = 11),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 9),
    panel.spacing = unit(1.2, "lines"),
    plot.margin = margin(t = 8, r = 12, b = 8, l = 12)
  )
```

Figure 4 compares homeownership rates with several key predictors using scatterplots and fitted trend lines. The clearest negative relationship appears between rent burden and homeownership: as rent burden increases, homeownership rates tend to decline. This supports the idea that states where renters face greater financial pressure may also be states where households have more difficulty transitioning into homeownership.

House price index, median income, and unemployment rate also show negative relationships with homeownership, though the strength and interpretation of these patterns vary. The negative relationship between house prices and homeownership is consistent with an affordability issue, since higher housing costs may make ownership less accessible. The median income relationship is more complicated because higher-income states may also be states with substantially higher housing costs, meaning income alone does not necessarily translate into easier access to homeownership. Mortgage rates and housing permits per person show slightly positive trend lines in this figure, but these relationships should be interpreted cautiously because they do not account for differences across states, changes over time, or other overlapping economic conditions.


```{r figure_5, echo=FALSE, message=FALSE, }
# Create box plot for homeownership rate by state
merged %>%
  ggplot(aes(
    x = reorder(state, homeownership_rate, median, na.rm = TRUE),
    y = homeownership_rate
  )) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Homeownership Rate by State",
    x = "State",
    y = "Homeownership Rate (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 6),
    plot.margin = margin(t = 8, r = 12, b = 8, l = 12)
  )
```

Figure 5 shows the distribution of homeownership rates by state across the study period. States are ordered by their typical homeownership rate, making it easy to identify states that consistently have higher or lower ownership levels. West Virginia, Michigan, Delaware, Maine, and Minnesota appear near the top of the distribution, while New York, California, Nevada, Hawaii, and Rhode Island appear near the bottom.

The lower-homeownership states are especially important for HUD because many of them are large or high-cost housing markets where affordability challenges are likely more severe. For example, California and New York have relatively low homeownership rates, which is consistent with the broader finding that higher housing costs and rent burden are associated with lower ownership. The chart also shows that some states have wider ranges than others, meaning homeownership conditions changed more over time in certain states while remaining more stable in others.


### **Multicollinearity Check:**
```{r multicollinearity_check, echo=FALSE, message=FALSE, }
#Check for multicolinearity
vif_model <- lm(
  homeownership_rate ~ mortgage_rate + hpi + unemployment_rate +
    median_income + rent_burden + perm_per_person,
  data = merged
)

vif(vif_model)
```

The multicollinearity check suggests that the predictors can be included together in the regression model without serious multicollinearity concerns. All variance inflation factor values are below the common threshold of 5, and well below the stricter threshold of 10, meaning that none of the predictors appears to be excessively redundant with the others. The highest values are for house price index and median income, which makes sense because states with higher incomes often also have higher housing prices. However, these values are still moderate, so the model can reasonably estimate the relationship between each predictor and homeownership while controlling for the others.

```{r lag_variables, message=FALSE, warning=FALSE, include=FALSE}
#Create lag variables
merged <- merged %>%
  arrange(state, year) %>%
  group_by(state) %>%
  mutate(
    mortgage_rate_lag1 = lag(mortgage_rate, 1),
    mortgage_rate_lag2 = lag(mortgage_rate, 2),
    mortgage_rate_lag3 = lag(mortgage_rate, 3),
    hpi_lag1 = lag(hpi, 1),
    hpi_lag2 = lag(hpi, 2),
    hpi_lag3 = lag(hpi, 3)
  ) %>%
  ungroup()
```

## **Regression Analysis**
For directions on how to conduct multilinear regression and plot confidence intervals, [click here](https://www.geeksforgeeks.org/r-language/how-to-use-the-coeftest-function-in-r/).

### **Basic Model**
```{r regression_basic, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model <- lm(homeownership_rate ~ mortgage_rate + state + 
               as.factor(year), data = merged )

#Summarize model
stargazer(model, type="text", digits = 2, 
          title = "Table 1: Simple OLS Regression", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate"), 
          covariate.labels=c("Mortgage Rate"))

#Create coefficients and standard errors
coefficients <- coef(model)[c("mortgage_rate")]
std_errors <- sqrt(diag(vcov(model)))[c("mortgage_rate")]

#Create data needed to plot confidence intervals
plot_data <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error, ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels = c(mortgage_rate = "Mortgage Rate"))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

[Discussion of basic model.]

### **Model 1**
```{r regression_1, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model1 <- lm(homeownership_rate ~ mortgage_rate + hpi + median_income +
               unemployment_rate + rent_burden + perm_per_person + state + 
               as.factor(year), data = merged )

#Summarize model
stargazer(model1, type="text", digits = 2, 
          title = "Table 2: OLS Regression with Controls", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "House Price Index", "Median Income", "Unemployment Rate",
          "Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model1)[c("mortgage_rate", "hpi", "median_income", 
                               "unemployment_rate", "rent_burden")]
std_errors <- sqrt(diag(vcov(model1)))[c("mortgage_rate", "hpi", 
                                         "median_income", "unemployment_rate", 
                                         "rent_burden")]

#Create data needed to plot confidence intervals
plot_data1 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data1, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels = c(
    hpi = "House Price Index",
    median_income = "Median Income",
    mortgage_rate = "Mortgage Rate",
    rent_burden = "Rent Burden",
    unemployment_rate = "Unemployment Rate"
  ))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

```{r clustered_standarderrors, echo=FALSE, message=FALSE, }
#Calculate clustered standard errors for Model 1
model1_cluster_se <- vcovCL(model1, cluster = ~ state)
summary_clustered <- coeftest(model1, vcov = model1_cluster_se)


```

[Discussion of model 1.]

### **Model 2**
```{r regression_2, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model2 <- lm(homeownership_rate ~ mortgage_rate + mortgage_rate_lag1 + 
               mortgage_rate_lag2 + mortgage_rate_lag3 + hpi + median_income +
               unemployment_rate + rent_burden + perm_per_person + 
               state + as.factor(year), data = merged )

#Summarize model
stargazer(model2, type="text", digits = 2, 
          title = "Table 3: OLS Regression with 3 Lags", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", "mortgage_rate_lag2", "mortgage_rate_lag3", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)", "Lagged Mortgage Rate (2)",
          "Lagged Mortgage Rate (3)", "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model2)[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "mortgage_rate_lag3", "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]
std_errors <- sqrt(diag(vcov(model2)))[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "mortgage_rate_lag3", "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]

#Create data needed to plot confidence intervals
plot_data2 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data2, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels = c("hpi" = "House Price Index",
                   "median_income" = "Median Income",
                   "mortgage_rate"= "Mortgage Rate", 
                   "mortgage_rate_lag1"= "Lagged Mortgage Rate (1)",
                   "mortgage_rate_lag2"= "Lagged Mortgage Rate (2)",
                   "mortgage_rate_lag3"= "Lagged Mortgage Rate (3)",
                   "rent_burden" = "Rent Burden",
                   "unemployment_rate" = "Unemployment Rate"))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

[Discussion of model 2.]

### **Model 3**
```{r regression_3, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model3 <- lm(homeownership_rate ~ mortgage_rate + mortgage_rate_lag1 + 
               mortgage_rate_lag2 + hpi + median_income +
               unemployment_rate + rent_burden + perm_per_person + state +
               as.factor(year), data = merged )

#Summarize model
stargazer(model2, type="text", digits = 2, 
          title = "Table 4: OLS Regression with 2 Lags", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", "mortgage_rate_lag2", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)", "Lagged Mortgage Rate (2)",
          "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model3)[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]
std_errors <- sqrt(diag(vcov(model3)))[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]

#Create data needed to plot confidence intervals
plot_data3 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data3, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  theme_minimal() +
  scale_x_discrete(labels=c("hpi" = "House Price Index",
                   "median_income" = "Median Income",
                   "mortgage_rate"= "Mortgage Rate", 
                   "mortgage_rate_lag1"= "Lagged Mortgage Rate (1)",
                   "mortgage_rate_lag2"= "Lagged Mortgage Rate (2)",
                   "rent_burden" = "Rent Burden",
                   "unemployment_rate" = "Unemployment Rate"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

[Discussion of model 3.]

### **Model 4**
```{r regression_4, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model4 <- lm(homeownership_rate ~ mortgage_rate + mortgage_rate_lag1 + 
               hpi + median_income + unemployment_rate + rent_burden + 
               perm_per_person + state + as.factor(year), data = merged )

#Summarize model
stargazer(model2, type="text", digits = 2, 
          title = "Table 5: OLS Regression with 1 Lag", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)",
          "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model4)[c("mortgage_rate", "mortgage_rate_lag1", "hpi", 
                               "median_income", "unemployment_rate",
                               "rent_burden")]
std_errors <- sqrt(diag(vcov(model4)))[c("mortgage_rate", "mortgage_rate_lag1", "hpi", "median_income", "unemployment_rate", "rent_burden")]

#Create data needed to plot confidence intervals
plot_data4 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data4, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels=c("hpi" = "House Price Index",
                   "median_income" = "Median Income",
                   "mortgage_rate"= "Mortgage Rate", 
                   "mortgage_rate_lag1"= "Lagged Mortgage Rate (1)",
                   "rent_burden" = "Rent Burden",
                   "unemployment_rate" = "Unemployment Rate"))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

[Discussion of model 4.]

### **Compare Models**
```{r regression_comparison, echo=FALSE, message=FALSE, warning = FALSE}
#Compare regression models using stargazer
stargazer(model1,model2,model3,model4, type="text", digits = 2, 
          title = "Table 6: Regression Comparisons)", style = "qje",
          column.labels = c('OLS', "OLS - 3 Lags", "OLS - 2 Lags", "OLS - 1 Lag"),
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", "mortgage_rate_lag2", "mortgage_rate_lag3", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)", "Lagged Mortgage Rate (2)",
          "Lagged Mortgage Rate (3)", "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

```

[Comparison of models and selection of model 4 as preffered model.]

# Conclusion

This project examined how mortgage interest rates, housing prices, and related economic and housing market variables are associated with homeownership rates across U.S. states over time. To answer this question, we constructed a state-year panel dataset combining homeownership rates, median income, population, rent burden, mortgage rates, house price index values, unemployment rates, and housing permits per person. We used descriptive visualizations, correlation analysis, state-level distribution plots, multicollinearity checks, and regression models to study both broad trends and more specific relationships among these variables.

The analysis shows that homeownership rates vary meaningfully across both time and states. Average homeownership declined from the mid-2000s into the mid-2010s before partially recovering in the late 2010s and early 2020s. However, this recovery occurred alongside a sharp rise in house price index values and, after 2021, a rapid increase in mortgage rates. This suggests that the current housing environment is shaped by multiple affordability pressures at once: households face not only higher home prices, but also higher borrowing costs.

Several patterns from the visual analysis help clarify the broader story. The correlation heatmap and scatterplots suggest that homeownership tends to be lower in states with higher population, higher rent burden, higher house price index values, and higher unemployment. Rent burden appears especially important because it connects the rental and ownership sides of the housing market: states where renters are already financially strained may also be states where households have more difficulty moving into ownership. The box plot by state also shows substantial geographic variation, with states such as New York, California, Nevada, Hawaii, and Rhode Island consistently showing lower homeownership rates than many other states.

The regression analysis adds further support to the idea that homeownership is influenced by several overlapping economic and housing market conditions. Mortgage rates are generally negatively associated with homeownership, especially in the baseline and controlled models, which is consistent with the idea that higher borrowing costs reduce access to ownership. Housing supply, measured through housing permits per person, appears positively associated with homeownership in the models, suggesting that states with more residential construction activity may have stronger ownership outcomes. Rent burden and unemployment also provide important context, although some relationships are more complex once controls and lagged variables are included.

There are several limitations to this analysis. First, the project uses observational data, so the results should be interpreted as associations rather than definitive causal effects. Second, some variables are measured at the national level, such as the 30-year mortgage rate, while others vary by state and year, which limits how precisely we can isolate state-specific mortgage effects. Third, the exclusion of 2020 due to ACS comparability issues helps maintain consistency but removes an important year from the housing market timeline. Finally, our housing supply measure captures permits rather than completed housing units, so it may not fully reflect the actual availability of homes.

Future work could improve the analysis by incorporating more localized data at the county or metropolitan level, adding measures of housing inventory and new construction completions, and exploring differences by household income, race, age, or first-time buyer status. Additional modeling could also test fixed effects more deeply, examine regional differences, or use causal inference methods to better isolate the effects of policy-relevant housing market changes.

# Policy Recommendation

Turning to policy, it should first and foremost be the priority to increase the housing supply, given the robust finding that more permits per capita result in higher rates of homeownership. To do so, certain barriers can be reduced. For example, working to update zoning regulations to streamline the use of land and permit more density and innovative construction techniques/housing types. It is important not simply to increase housing generally but to increase affordable housing, as homeownership rates are lower in largely populated areas where the rent burden is already high. In addition to increasing the homeownership rate, increasing access to affordable housing also has a wealth of other benefits for people. There have been several initiatives at the state level that can be applied at the federal level to improve homeownership generally, as well as to help states that are at a lower rate potentially catch up. Actions like lessening parking requirements, density limits, and minimum lot sizes for new residential buildings allow for more financially feasible constructions. Thus, finding ways to reduce these kinds of restrictive zoning, which are otherwise contributing to the affordability crisis in the US, may further aid this endeavor to increase homeownership rates.

With high rent burden correlated with low homeownership, actions should be taken to also address this renter issue to allow for mobility toward homeownership for those who want to purchase a house but have historically financially struggled to do so because of the already large burden of rent. This could be approached by the development of more affordable rental properties and/or expanding rental assistance.  We also need to turn to the impact mortgage rates have, with a one percentage point increase resulting in both an immediate and long-term negative effect on homeownership. With mortgage rates rising and staying high in recent years, it may be helpful to expand access to and opportunity for fixed-rate mortgages at lower rates. Combined, further resources and energy need to be directed toward the HOME Investment Partnerships Program, both at the rental (renter and rental construction development) and homeownership levels.

Specific focus should be given to states with the lowest rates of homeownership, like California, New York, Nevada, and Hawaii. Those at the federal level should aim to work with and hear from homeowners, developers, and local and state legislators to best understand the needs of those in areas struggling most. Inspiration can also be taken from successful state policies, especially in states with higher rates of homeownership, more permits per person, and lower rent burden.


# Recorded Policy Brief

TO BE ADDED BY MAY 11.

# Version Control

[Version control text here.]

# References

Federal Housing Finance Agency. (n.d.). *FHFA House Price Index® datasets*. Retrieved 2026, from https://www.fhfa.gov/data/house-price-index

Federal Reserve Bank of St. Louis. (n.d.). *30-year fixed rate mortgage average in the United States [MORTGAGE30US]*. FRED. Retrieved 2026, from https://fred.stlouisfed.org/series/MORTGAGE30US

U.S. Bureau of Labor Statistics. (n.d.). *Local Area Unemployment Statistics*. Retrieved 2026, from https://www.bls.gov/lau/

U.S. Census Bureau. (n.d.). *American Community Survey data via API*. Retrieved 2026, from https://www.census.gov/programs-surveys/acs/data/data-via-api.html

U.S. Census Bureau. (n.d.). *Building Permits Survey*. Retrieved 2026, from https://www.census.gov/permits
