---
title: "Class Project"
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document: default
---


### **This landing page displays the knitted output of our README.Rmd file. For the code behind the analysis and figures shown below, please consult the README.Rmd file.**

# Repository Guide

This repository contains our group project for Software Tools for Data Analysis. Our project studies how mortgage interest rates and housing prices affect homeownership rates across U.S. states over time. The final audience for this work is Scott Turner, U.S. Secretary of Housing and Urban Development (HUD).

### Branch Structure
- `main`: current working branch for the final project deliverables
- `Checkpoint-1` branch: preserved as a historical archive of our first checkpoint submission
- `Checkpoint-2` branch: preserved as a historical archive of our second checkpoint submission

### Repository Contents
- `README.Rmd`: source file for this repository landing page
- `README.md`: knitted GitHub-facing version of the README
- `Data/`: raw and cleaned data files used in the analysis
- `Report Outline.docx`: working outline for the written report
- `README_files/`: figures generated from the README


```{r clear_environment, include=FALSE}
remove(list=ls())
```

# Introduction


Homeownership remains one of the most important pathways to wealth-building, stability, and long-term financial security in the United States. However, recent housing market conditions have made homeownership increasingly difficult to attain for many families and individuals. Higher mortgage rates raise the cost of borrowing, while rising housing prices increase the upfront and long-term financial burden of purchasing a home. These pressures are especially important for the U.S. Department of Housing and Urban Development (HUD), which is responsible for supporting fair, affordable, and sustainable access to housing across the country.

This report is written for Scott Turner, U.S. Secretary of Housing and Urban Development. As HUD considers how to address challenges with affordability, expand access to homeownership, and support stable housing markets, policymakers need evidence on which factors are most closely associated with changes in homeownership rates. While mortgage rates and housing prices are central to the current affordability discussion, homeownership is also shaped by broader state-level conditions. Some of these include income, unemployment, rent burden, and housing supply. For that reason, our analysis examines both the direct relationship between borrowing costs, housing prices, and homeownership, as well as the role of additional economic and housing market drivers.

Our research question is: **How do mortgage interest rates, housing prices, and related state-level economic and housing market conditions affect homeownership rates across the United States over time?** To answer this question, we construct a state-year panel dataset covering U.S. states from 2005 to 2024. We exclude 2020 due to comparability issues in one-year American Community Survey data during the COVID-19 period. We combine public data from the U.S. Census Bureau, Federal Reserve Economic Data (FRED), the Federal Housing Finance Agency (FHFA), the Bureau of Labor Statistics (BLS), and the Census Building Permits Survey.

The results of this analysis can help HUD evaluate which policy levers may be most relevant for improving access to homeownership. If homeownership rates are strongly related to mortgage rates, housing prices, rent burden, or housing supply, then federal housing policies can be better targeted toward fixing conditions that most constrain citizens from purchasing homes. These decisions have consequences beyond individual buyers: homeownership patterns affect household wealth accumulation, neighborhood stability, housing market equity, and the broader economic health of communities.


# Data Summary

To study these relationships, we construct a state-year panel dataset covering U.S. states from 2005 to 2024, excluding 2020 because one-year American Community Survey data were not available in a comparable form during the COVID-19 period. Each observation in the dataset represents one state in one year. This structure allows us to compare both differences across states and changes within states over time, making it well suited for analyzing how economic and housing market conditions are associated with homeownership outcomes.

The dataset combines several reputable public sources. Homeownership rates, median income, population, and rent burden are drawn from the U.S. Census Bureau’s American Community Survey. Mortgage rates come from the Federal Reserve Economic Data database, while state-level housing price data come from the Federal Housing Finance Agency House Price Index. We also include unemployment data from the Bureau of Labor Statistics and housing permit data from the Census Building Permits Survey as a proxy for housing supply. After importing, cleaning, and merging these sources, we use the resulting dataset to examine patterns in homeownership, affordability, labor market conditions, and housing supply across states. This combined dataset provides the foundation for both our descriptive analysis and our regression modeling.


```{r load_packages, message=FALSE, warning=FALSE, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))
# Load required packages
packages <- c(
  # data manipulation
  "tidyverse",
  "janitor",
  # data import/extraction
  "readr",
  "readxl",
  "car",
  "tidycensus",
  "fredr",
  # dates and time handling
  "lubridate",
  # time series structure and analysis
  "tsibble",
  "fpp3",
  "patchwork",
  # regression
  "sandwich",
  "lmtest",
  "stargazer",
  "modelsummary"
)

# checks if package available (if not - install packages)
# load packages
check.packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}
check.packages(packages)
```


```{r set_api_keys, include=FALSE}
### **Set API Keys:**
#tidycensus api key: https://api.census.gov/data/key_signup.html
#fredr api key: https://fred.stlouisfed.org/docs/api/api_key.html

census_api_key("894b3464183cec8f6c0fa053eb34d0d878de682e")
fredr_set_key("8351bf4f459d0e8c92d56f50021947d8")
```


```{r import_data, message=FALSE, warning=FALSE, include=FALSE}
#Set years to include the last two decades, excluding 2020 due to the COVID-19 pandemic.
years <- c(2005:2019, 2021:2024)

#Extract home ownership data from the Census
homeownership <- map_dfr(
  years,
  function(yr) {
    get_acs(
      geography = "state",
      variables = c(
        total_occupied = "B25003_001",
        owner_occupied = "B25003_002"
      ),
      year = yr,
      survey = "acs1"
    ) %>%
      select(NAME, variable, estimate) %>%
      pivot_wider(names_from = variable, values_from = estimate) %>%
      mutate(
        year = yr,
        homeownership_rate = 100 * owner_occupied / total_occupied
      ) %>%
      transmute(
        state = NAME,
        year,
        homeownership_rate
      )
  }
)

#Extract income data from the Census
income_list <- map(
  years,
  ~ get_acs(
      geography = "state",
      variables = "B19013_001",
      year = .x,
      survey = "acs1"
    ) %>%
      transmute(
        state = NAME,
        year = .x,
        median_income = estimate
      )
)

#Bind the income data into a DF
income <- bind_rows(income_list)

#Extract population data from the Census
population_list <- map(
  years,
  ~ get_acs(
      geography = "state",
      variables = "B01003_001",
      year = .x,
      survey = "acs1"
    ) %>%
      transmute(
        state = NAME,
        year = .x,
        population = estimate
      )
)

#Bind the population data into a DF
population <- bind_rows(population_list)

# Rent burden from ACS table B25070
# Defined as share of renter households paying 30% or more of income toward gross rent
rent_burden <- map_dfr(
  years,
  function(yr) {
    get_acs(
      geography = "state",
      table = "B25070",
      year = yr,
      survey = "acs1"
    ) %>%
      select(NAME, variable, estimate) %>%
      mutate(year = yr) %>%
      pivot_wider(names_from = variable, values_from = estimate) %>%
      transmute(
        state = NAME,
        year,
        rent_burden = 100 * (
          B25070_007 + B25070_008 + B25070_009 + B25070_010
        ) / B25070_001
      )
  }
)

# Housing supply from Census Building Permits Survey
# Read directly from Census-hosted Excel file
permits_url <- "https://www.census.gov/construction/bps/xls/annualhistorybystate.xls"

tmp_permits <- tempfile(fileext = ".xls")
download.file(permits_url, tmp_permits, mode = "wb")

sheet_names <- excel_sheets(tmp_permits)
sheet_names <- sheet_names[sheet_names != "PR"]

permits_state <- map_dfr(
  sheet_names,
  function(sh) {
    read_excel(
      tmp_permits,
      sheet = sh,
      skip = 5,
      col_names = FALSE
    ) %>%
      select(1:5) %>%
      setNames(c(
        "universe",
        "year",
        "total_buildings",
        "housing_permits",
        "total_valuation"
      )) %>%
      filter(
        !is.na(year),
        str_detect(as.character(year), "^[0-9]{4}$")
      ) %>%
      transmute(
        state = state.name[match(sh, state.abb)],
        year = as.integer(year),
        housing_permits = readr::parse_number(as.character(housing_permits))
      ) %>%
      filter(year %in% years)
  }
)

#Extract mortgage data from FRED
mortgage_raw <- fredr(
  series_id = "MORTGAGE30US",
  observation_start = as.Date("2005-01-01"),
  observation_end = Sys.Date()
)

#Estimate annual mortgage rates
mortgage_annual <- mortgage_raw %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarise(
    mortgage_rate = mean(value, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r import_local_data, message=FALSE, warning=FALSE, include=FALSE}
# Import Local Data
fhfa_raw <- read_csv("Data/hpi_at_state (1).csv", col_names = FALSE) %>%
  setNames(c("state", "year", "quarter", "hpi"))

# Set HPI data by state/year
fhfa_annual <- fhfa_raw %>%
  clean_names() %>%
  filter(year >= 2005) %>%
  group_by(state, year) %>%
  summarise(
    hpi = mean(hpi, na.rm = TRUE),
    .groups = "drop"
  )

hpi_annual <- fhfa_annual %>%
  mutate(state = state.name[match(state, state.abb)])

# Import local unemployment data
bls_data <- read_delim("Data/la.data.2.AllStatesU.txt", delim = "\t") %>%
  clean_names()

bls_series <- read_delim("Data/la.series.txt", delim = "\t") %>%
  clean_names()

bls_area <- read_delim("Data/la.area", delim = "\t") %>%
  clean_names()

# Join the actual data to the series
bls_merged <- bls_data %>%
  left_join(bls_series, by = "series_id")

# Use key to select code
bls_merged <- bls_merged %>%
  left_join(
    bls_area %>% select(area_code, area_text),
    by = "area_code"
  )

# Keep annual-average unemployment rate
state_unemployment <- bls_merged %>%
  filter(period == "M13") %>%          # annual average
  filter(measure_code == "03") %>%     # unemployment rate
  transmute(
    state = area_text,
    year = as.integer(year),
    unemployment_rate = as.numeric(value)
  ) %>%
  filter(state %in% c(state.name, "District of Columbia")) %>%
  arrange(state, year)
```



```{r merge_data, message=FALSE, warning=FALSE, include=FALSE}
#Merge the data sets by year and state
merged <- left_join(
  population,
  income,
  by = c("state", "year")
) %>%
  left_join(
    homeownership,
    by = c("state", "year")
  ) %>%
  left_join(
    hpi_annual,
    by = c("state", "year")
  ) %>%
  left_join(
    state_unemployment,
    by = c("state", "year")
  ) %>%
  left_join(
    rent_burden,
    by = c("state", "year")
  ) %>%
  left_join(
    permits_state,
    by = c("state", "year")
  ) %>%
  left_join(
    mortgage_annual,
    by = "year"
  )
```


```{r permits_per_person, message=FALSE, warning=FALSE, include=FALSE}
#Estimate annual mortgage rates
merged <- merged %>%
  mutate(perm_per_person = housing_permits/population, NA)
```


```{r excludeDCandPR, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Filter out D.C. and PR
merged <- merged %>%
  filter(!state %in% c("Puerto Rico", "District of Columbia"))

#Remove a bad column
merged <- merged[, names(merged) != "NA"]
```

### **Summary Statistics:**
```{r pop_summary_stats, echo=FALSE, warning=FALSE, message=FALSE}
#Make a df version of merged
merged_df <- as.data.frame(merged)

#Rename quantitative variables in merged_df
merged_df <- merged_df %>%
       select('Population' = population,
              "Median Income" = median_income,
              "Homeownership Rate" = homeownership_rate,
              "Mortgage Rate" = mortgage_rate,
              "Housing Price Index" = hpi,
              "Unemployment Rate" = unemployment_rate,
              "Rent Burden" = rent_burden,
              "Housing Permits" = housing_permits,
              "Housing Permits Per Person" = perm_per_person)

#Compute summary statistics for all quantitative variables
datasummary_skim(merged_df, histogram=FALSE, title = "Summary Statistics", align="c")

```


The summary statistics above provide information about central tendencies and variation across key variables over the period between 2005 and 2024 (excluding 2020). Note: these statistics were calculated treating each state/year combination as a single observation. Statistics reflect state conditions over this period and are not weighted by population.

The average state had a population of 6.3 million, a median income of approximately $58,000, and an unemployment rate of 5.3 percent over the analysis timeframe. The average state homeownership rate was 67 percent, while the average national mortgage rate was 4.86 percent. An average of just 0.004 housing permits were issued on annual basis per person, and 44 percent of the state population was rent burdened, on average.

The homeownership rate ranged from 52 percent to 76 percent, with a standard deviation of 4.4 percent. The fact that the average homeownership rate (67.0 percent) and median homeownership rate (67.5 percent) were similar, suggests that the distribution is roughly normal (a histogram and accompanying analysis are provided below).

Several potential drivers of the homeownership rate showed substantial variation across the analysis timeframe. The mortgage rate ranged from 3 to 6.8 percent with a standard deviation of 1.2 percent; the housing price index ranged from 184to 1,230, with a standard deviation of 164; and rent burden ranged from 28.3 to 58.1 percent with a standard deviation of 4.4 percent. The variation of these variables in our sample will allow us to meaningfully test their influence on the homeownership rate.



### **Plot Histograms of Key Variables:**
```{r figure_1, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=7.5}
key_vars_labs <- c("homeownership_rate" = "Homeownership Rate", "hpi" = "House Price Index", "median_income" = "Median Income", "population" = "Population", "rent_burden" = "Rent Burden", "unemployment_rate" = "Unemployment Rate", "perm_per_person" = "Housing Permits per Person")

#Create histogram plots of key variables
merged %>%
  select(
    homeownership_rate,
    median_income,
    population,
    hpi,
    unemployment_rate,
    rent_burden,
    perm_per_person
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 2, 
             labeller = as_labeller(key_vars_labs)) +
  labs(
    x = NULL,
    y = "Count"
  ) +
  theme_minimal()
```

The histograms shown above provide useful information about the distribution of homeownership rate and factors that may influence it. The histograms were generally unimodal and bell shaped, though several of them revealed significant kurtosis. Consistent with the observation above that the mean and median homeownership rate are similar, the histogram for the homeownership rate shows a mostly normal distribution, albeit slightly skewed to the left. We see most states had homeownership rates between 60 and 75 percent over the analysis timeframe. Only a handful had homeownership rates below 60 percent. Rent burden data appears normally distributed. With data distributed normally like this, the mean can be a reliable representation for an average data point. Since homeownership rate data is slightly left-skewed, its mean might be slightly lower than where the graph actually peaks.

The housing price index and median income variables were both heavily skewed to the right, with most median income values between $40,000 and $70,000. This distribution of the data demonstrates that only a handful of states were significantly wealthier or had higher housing prices than their counterparts. 

Housing permits per person also had a noticeable righthand tail, with the majority of values being generally quite low. This suggests that a few states were issuing permits faster (on a per person basis) a significantly faster than other states and that it is infrequent to generally see high levels of permits given that even the outliers were relatively low in count. 

The fact that population and the unemployment rate were also both right-skewed is as expected. A handful of states (California, Texas, Florida, and New York) are significantly more populous than the average state, and the unemployment rate can more than double its typical level during economic recessions.


# Data Analytics

### **Key Trends Over Time:**

```{r figure_2, echo=FALSE, message=FALSE,  warning=FALSE,}
#| fig-width: 10
#| fig-height: 7
# Average homeownership rate over time
avg_homeownership_ts <- merged %>%
  group_by(year) %>%
  summarise(
    avg_homeownership_rate = mean(homeownership_rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as_tsibble(index = year)

p1 <- autoplot(avg_homeownership_ts, avg_homeownership_rate) +
  labs(
    title = "Average State Homeownership Rate Over Time",
    x = "Year",
    y = "Homeownership Rate (%)"
  ) +
  theme(
    plot.title = element_text(size = 18),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.margin = margin(t = 6, r = 10, b = 6, l = 10)
  )

# Average HPI over time
avg_hpi_ts <- merged %>%
  group_by(year) %>%
  summarise(
    avg_hpi = mean(hpi, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as_tsibble(index = year)

p2 <- autoplot(avg_hpi_ts, avg_hpi) +
  labs(
    title = "Average State House Price Index Over Time",
    x = "Year",
    y = "House Price Index"
  ) +
  theme(
    plot.title = element_text(size = 18),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.margin = margin(t = 6, r = 10, b = 6, l = 10)
  )

# Mortgage rate over time
mortgage_ts <- mortgage_annual %>%
  as_tsibble(index = year)

p3 <- autoplot(mortgage_ts, mortgage_rate) +
  labs(
    title = "Average 30-Year Mortgage Rate Over Time",
    x = "Year",
    y = "Mortgage Rate (%)"
  ) +
  theme(
    plot.title = element_text(size = 18),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.margin = margin(t = 6, r = 10, b = 6, l = 10)
  )

# Combine all three plots
p1 / p2 / p3
```
Figure 2 shows broad time trends in average state homeownership rates, house price index values, and 30-year mortgage rates. 

Exploring average state homeownership across time illustrates that it peaked right before and declined following the 2008 recession. It hit its minimum between 2014 and 2016 before gradually recovering in the late 2010s and early 2020s, though not yet back to the levels seen in 2005. Over the same period, the average state HPI (House Price Index) increased substantially, suggesting that housing prices rose much faster than homeownership rates recovered. The HPI, a means of measuring the rate at which home prices are changing across the housing market in the U.S., is an important indicator of the housing market as it reflects the kind of market (whether buyers or sellers are more so advantaged), and the year-to-year change indicating strength and growth. The mortgage rate trend provides additional context. The 30-year mortgage rate started to decline around and following the 2008 recession, staying consistently low from about 2012 through 2016. While it once again plummeted from roughly 2018 through 2021, it saw a sharp rise immediately after and only recently started to plateau – at its highest level in the past 20 years. This creates an important policy concern, as even though homeownership partially recovered after the mid-2010s, households now face a combination of elevated housing prices and higher borrowing costs. Together, these trends suggest that homeownership affordability is shaped by multiple pressures at once rather than by any single variable alone.

### **Correlation Matrix:**
```{r figure_3, echo=FALSE, message=FALSE}
#Create correlation matrix
#Select variables of interest
corr_data <- merged %>%
  select(
    homeownership_rate,
    median_income,
    population,
    hpi,
    unemployment_rate,
    mortgage_rate,
    rent_burden,
    perm_per_person
  )

corr_matrix <- cor(corr_data, use = "complete.obs")

corr_df <- as.data.frame(as.table(corr_matrix))

# labels
var_labels <- c(
  homeownership_rate = "Homeownership Rate",
  median_income      = "Median Income",
  population         = "Population",
  hpi                = "House Price Index",
  unemployment_rate  = "Unemployment Rate",
  mortgage_rate      = "Mortgage Rate",
  rent_burden        = "Rent Burden",
  perm_per_person    = "Housing Permits per Person"
)

# keep only upper triangle, relabel
corr_df <- corr_df %>%
  mutate(
    Var1 = factor(Var1, levels = colnames(corr_matrix)),
    Var2 = factor(Var2, levels = colnames(corr_matrix))
  ) %>%
  filter(as.numeric(Var1) <= as.numeric(Var2)) %>%
  mutate(
    Var1 = factor(var_labels[as.character(Var1)], levels = var_labels),
    Var2 = factor(var_labels[as.character(Var2)], levels = var_labels)
  )

#Plot the correlation matrix
ggplot(corr_df, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", Freq)), size = 3.5) +
  scale_fill_gradient2(
    low = "firebrick",
    mid = "white",
    high = "steelblue",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation Heatmap of Key Variables",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

Figure 3, the correlation heatmap provides an initial look at the relationships among different variables before moving into regression analysis. Homeownership has negative correlations with population, rent burden, house price index, median income, and unemployment, indicates that states with larger populations, higher housing costs, and greater rental affordability pressure tend to have lower homeownership rates. Homeownership rate’s moderately negative correlation with population and rent burden may also suggest that since areas with higher populations had lower rates of homeownership there were likely higher levels of renting. The negative relationship with rent burden is especially important because it suggests that places where renters face greater financial strain may also be places where households have more difficulty transitioning into ownership.

The heatmap also shows several relationships among the explanatory variables themselves. Median income and house price index have a strong positive correlation, indicating that higher-income states also tend to have more expensive housing markets. Rent burden (spending more than 30% of gross household income on rent; high rental cost relative to income) is moderately, positively correlated with population and house price index. In other words, where the population is larger, there is a larger rent burden; this supports the broader affordability story that larger and more expensive states tend to place more pressure on renters. These correlations help motivate the use of multivariable regression, since homeownership is related to several overlapping economic and housing market conditions rather than one factor in isolation.


```{r missing_check, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Check for missing obsevations
merged %>%
  summarise(
    across(
      c(
        homeownership_rate,
        median_income,
        population,
        hpi,
        unemployment_rate,
        mortgage_rate,
        rent_burden,
        perm_per_person
      ),
      ~ sum(is.na(.))
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  )
```


```{r figure_4, echo=FALSE, message=FALSE, }
# Create lines of best fit for key variables

# Identify key variables
kpi_labs <- c(
  "hpi" = "House Price Index",
  "median_income" = "Median Income", 
  "mortgage_rate" = "Mortgage Rate", 
  "perm_per_person" = "Housing Permits per Person", 
  "rent_burden" = "Rent Burden", 
  "unemployment_rate" = "Unemployment Rate"
)

# Plot the lines of best fit
merged %>%
  select(
    homeownership_rate,
    mortgage_rate,
    hpi,
    unemployment_rate,
    median_income,
    rent_burden,
    perm_per_person
  ) %>%
  pivot_longer(
    cols = -homeownership_rate,
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value, y = homeownership_rate)) +
  geom_point(alpha = 0.35, size = 1.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 0.9) +
  facet_wrap(
    ~ variable,
    scales = "free_x",
    ncol = 2,
    labeller = as_labeller(kpi_labs)
  ) +
  labs(
    title = "Homeownership Rate vs. Key Predictors",
    x = NULL,
    y = "Homeownership Rate (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16),
    strip.text = element_text(size = 11),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 9),
    panel.spacing = unit(1.2, "lines"),
    plot.margin = margin(t = 8, r = 12, b = 8, l = 12)
  )
```

Figure 4 compares homeownership rates with several key predictors using scatterplots and fitted trend lines. The clearest negative relationship appears between rent burden and homeownership: as rent burden increases, homeownership rates tend to decline. This supports the idea that states where renters face greater financial pressure may also be states where households have more difficulty transitioning into homeownership.

House price index, median income, and unemployment rate also show negative relationships with homeownership, though the strength and interpretation of these patterns vary. The negative relationship between house prices and homeownership is consistent with an affordability issue, since higher housing costs may make ownership less accessible. The median income relationship is more complicated because higher-income states may also be states with substantially higher housing costs, meaning income alone does not necessarily translate into easier access to homeownership. Mortgage rates and housing permits per person show slightly positive trend lines in this figure, but these relationships should be interpreted cautiously because they do not account for differences across states, changes over time, or other overlapping economic conditions.


```{r figure_5, echo=FALSE, message=FALSE, }
# Create box plot for homeownership rate by state
merged %>%
  ggplot(aes(
    x = reorder(state, homeownership_rate, median, na.rm = TRUE),
    y = homeownership_rate
  )) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Homeownership Rate by State",
    x = "State",
    y = "Homeownership Rate (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 6),
    plot.margin = margin(t = 8, r = 12, b = 8, l = 12)
  )
```

Figure 5 shows the distribution of homeownership rates by state across the study period. States are ordered by their typical homeownership rate, making it easy to identify states that consistently have higher or lower ownership levels. West Virginia, Michigan, Delaware, Maine, and Minnesota appear near the top of the distribution (~ 70-75% homeownership rate), while New York, California, Nevada, Hawaii, and Rhode Island (~ 55-60% homeownership rate) appear near the bottom. 

The lower-homeownership states are especially important for HUD because many of them are large or high-cost housing markets where affordability challenges are likely more severe. For example, California and New York have relatively low homeownership rates, which is consistent with the broader finding that higher housing costs and rent burden are associated with lower ownership. The chart also shows that some states have wider ranges than others, meaning homeownership conditions changed more over time in certain states while remaining more stable in others.


### **Multicollinearity Check:**
```{r multicollinearity_check, echo=FALSE, message=FALSE, }
#Check for multicolinearity
vif_model <- lm(
  homeownership_rate ~ mortgage_rate + hpi + unemployment_rate +
    median_income + rent_burden + perm_per_person,
  data = merged
)

vif(vif_model)
```

The multicollinearity check suggests that the predictors can be included together in the regression model without serious multicollinearity concerns. All variance inflation factor values are below the common threshold of 5, and well below the stricter threshold of 10, meaning that none of the predictors appears to be excessively redundant with the others. The highest values are for house price index and median income, which makes sense because states with higher incomes often also have higher housing prices. However, these values are still moderate, so the model can reasonably estimate the relationship between each predictor and homeownership while controlling for the others.

```{r lag_variables, message=FALSE, warning=FALSE, include=FALSE}
#Create lag variables
merged <- merged %>%
  arrange(state, year) %>%
  group_by(state) %>%
  mutate(
    mortgage_rate_lag1 = lag(mortgage_rate, 1),
    mortgage_rate_lag2 = lag(mortgage_rate, 2),
    mortgage_rate_lag3 = lag(mortgage_rate, 3),
    hpi_lag1 = lag(hpi, 1),
    hpi_lag2 = lag(hpi, 2),
    hpi_lag3 = lag(hpi, 3)
  ) %>%
  ungroup()
```

## **Regression Analysis**
After exploring patterns in the data and potential relationships between key variables, we then engaged in more rigorous econometric modeling to test which variables meaningfully influence the homeownership rate after controlling for observable factors as well as state and year fixed effects. The results of that analysis are shown below.

For directions on how to conduct multilinear regression and plot confidence intervals, [click here](https://www.geeksforgeeks.org/r-language/how-to-use-the-coeftest-function-in-r/).

### **Model 1**
```{r regression_1, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model1 <- lm(homeownership_rate ~ mortgage_rate + hpi + median_income +
               unemployment_rate + rent_burden + perm_per_person + state + 
               as.factor(year), data = merged )

#Summarize model
stargazer(model1, type="text", digits = 2, 
          title = "Table 2: OLS Regression with Controls", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "House Price Index", "Median Income", "Unemployment Rate",
          "Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model1)[c("mortgage_rate", "hpi", "median_income", 
                               "unemployment_rate", "rent_burden")]
std_errors <- sqrt(diag(vcov(model1)))[c("mortgage_rate", "hpi", 
                                         "median_income", "unemployment_rate", 
                                         "rent_burden")]

#Create data needed to plot confidence intervals
plot_data1 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data1, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels = c(
    hpi = "House Price Index",
    median_income = "Median Income",
    mortgage_rate = "Mortgage Rate",
    rent_burden = "Rent Burden",
    unemployment_rate = "Unemployment Rate"
  ))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

```{r clustered_standarderrors, echo=FALSE, message=FALSE, }
#Calculate clustered standard errors for Model 1
model1_cluster_se <- vcovCL(model1, cluster = ~ state)
summary_clustered <- coeftest(model1, vcov = model1_cluster_se)


```

In the first model we estimated included all our variables of interest (mortgage rate, housing price index, rent burden, and housing permits per person) as well as control variables (median income and the unemployment rate) and state and year fixed effects. The R2 value of 0.97 tells us that 97 percent of the variability in the homeownership rate is explained by the parameters included in the model. The mortgage rate, housing price index, rent burden, and housing permits per person are all significant at the one percent level, while the unemployment rate is significant at the five percent level.

Several of the variables had effects consistent with what one would expect. For example, the coefficient on the mortgage rate implies that one percentage point increase is associated with a 2.14 percentage point reduction in the homeownership rate. A 0.1 increase in annual housing permits per person is associated with a 7.7 percentage point increase in homeownership rate, whereas a one percentage point decrease in the unemployment rate was associated with a 0.1 percentage reduction.

Other variables had coefficients that appeared potentially counter intuitive. For example, the coefficient on the housing price index was positive, implying that states with less affordable housing had higher homeownership rates, all else being equal. This is theoretically inconsistent and is suggestive of omitted variable bias (i.e., an unobserved variable may be correlated with the housing price index). It was also surprising to observe that rent burden had a positive coefficient, which implies that states with a greater share of the population burdened by rent had higher homeownership rates, all else being equal. One possible explanation is that individuals living in states with unusually high rent may have an incentive to buy their first home earlier in life rather than wait and save for their ideal home.

A key weakness of this model is that it does not account for the fact that the mortgage rate likely has a lagged effect on the homeownership rate. That is, individuals may observe a change in the mortgage rate, then start the process of saving for and financing a home, which can take months or years. Thus, it is possible that changes in the mortgage rate influences human behavior and the homeownership rate in the later years. To explore this relationship, we estimated three more regression models: one with three lagged mortgage rate variables, one with two lagged variables, and one with one lagged variable. The results of those models are shown below.

### **Model 2**
```{r regression_2, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model2 <- lm(homeownership_rate ~ mortgage_rate + mortgage_rate_lag1 + 
               mortgage_rate_lag2 + mortgage_rate_lag3 + hpi + median_income +
               unemployment_rate + rent_burden + perm_per_person + 
               state + as.factor(year), data = merged )

#Summarize model
stargazer(model2, type="text", digits = 2, 
          title = "Table 3: OLS Regression with 3 Lags", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", "mortgage_rate_lag2", "mortgage_rate_lag3", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)", "Lagged Mortgage Rate (2)",
          "Lagged Mortgage Rate (3)", "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model2)[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "mortgage_rate_lag3", "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]
std_errors <- sqrt(diag(vcov(model2)))[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "mortgage_rate_lag3", "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]

#Create data needed to plot confidence intervals
plot_data2 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data2, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels = c("hpi" = "House Price Index",
                   "median_income" = "Median Income",
                   "mortgage_rate"= "Mortgage Rate", 
                   "mortgage_rate_lag1"= "Lagged Mortgage Rate (1)",
                   "mortgage_rate_lag2"= "Lagged Mortgage Rate (2)",
                   "mortgage_rate_lag3"= "Lagged Mortgage Rate (3)",
                   "rent_burden" = "Rent Burden",
                   "unemployment_rate" = "Unemployment Rate"))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Including three lagged variables resulted in theoretically inconsistent results. The three lagged mortgage rate variables all had small positive effects on the homeownership rate, which is inconsistent and likely due to model noise.

### **Model 3**
```{r regression_3, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model3 <- lm(homeownership_rate ~ mortgage_rate + mortgage_rate_lag1 + 
               mortgage_rate_lag2 + hpi + median_income +
               unemployment_rate + rent_burden + perm_per_person + state +
               as.factor(year), data = merged )

#Summarize model
stargazer(model2, type="text", digits = 2, 
          title = "Table 4: OLS Regression with 2 Lags", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", "mortgage_rate_lag2", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)", "Lagged Mortgage Rate (2)",
          "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model3)[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]
std_errors <- sqrt(diag(vcov(model3)))[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]

#Create data needed to plot confidence intervals
plot_data3 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data3, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  theme_minimal() +
  scale_x_discrete(labels=c("hpi" = "House Price Index",
                   "median_income" = "Median Income",
                   "mortgage_rate"= "Mortgage Rate", 
                   "mortgage_rate_lag1"= "Lagged Mortgage Rate (1)",
                   "mortgage_rate_lag2"= "Lagged Mortgage Rate (2)",
                   "rent_burden" = "Rent Burden",
                   "unemployment_rate" = "Unemployment Rate"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The third model included two lagged mortgage rate variables. The results of this model were even more nonsensical. The mortgage rate variable and the two-year lagged variable both had large, negative coefficients, while the coefficient on the one year lagged variable was similarly large and positive. All three coefficient were significant at the one percent level.

### **Model 4**
```{r regression_4, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model4 <- lm(homeownership_rate ~ mortgage_rate + mortgage_rate_lag1 + 
               hpi + median_income + unemployment_rate + rent_burden + 
               perm_per_person + state + as.factor(year), data = merged )

#Summarize model
stargazer(model2, type="text", digits = 2, 
          title = "Table 5: OLS Regression with 1 Lag", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)",
          "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model4)[c("mortgage_rate", "mortgage_rate_lag1", "hpi", 
                               "median_income", "unemployment_rate",
                               "rent_burden")]
std_errors <- sqrt(diag(vcov(model4)))[c("mortgage_rate", "mortgage_rate_lag1", "hpi", "median_income", "unemployment_rate", "rent_burden")]

#Create data needed to plot confidence intervals
plot_data4 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data4, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels=c("hpi" = "House Price Index",
                   "median_income" = "Median Income",
                   "mortgage_rate"= "Mortgage Rate", 
                   "mortgage_rate_lag1"= "Lagged Mortgage Rate (1)",
                   "rent_burden" = "Rent Burden",
                   "unemployment_rate" = "Unemployment Rate"))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The fourth and final model included just one lagged variable. The coefficient on the one year lagged mortgage rate variable is modest, but negative as expected and statistically significant at the five percent level. The coefficients on the other variables are generally similar to the ones from the first model, with two notable exceptions. First the effect of the current mortgage rate is shown to be nearly three times as big. In model 4, a one percentage point increase in the current mortgage rate is associated with a 5.9 percentage point reduction in the homeownership rate, all else being equal (compared to the 2.1 percentage reduction shown in model 1). The second notable change was to the housing permits per person variable. In Model 4, a 0.1 increase in annual housing permits per person is associated with a 9.1 percentage point increase in homeownership rate (compared to the 7.7 percentage point reduction shown in Model 1).

Model 4 is our preferred model. We believe there is strong theoretical justification that the mortgage rate would have a delayed effect on the homeownership rate (i.e., saving and financing a house does not happen overnight). Including multiple lags disrupted the model in ways that undermined reason (coefficients were unintuitive, the effect of the unemployment rate was no longer statistically significant, etc.). However, including just one lagged mortgage rate variable seemed to improve the model. The coefficient on the lagged variable was negative as expected (and statistically significant), and the other variables were not influenced in implausible ways. As noted, the coefficients on the current mortgage rate and housing permits per person both increased in magnitude, but not to an unreasonable degree.

### **Compare Models**
```{r regression_comparison, echo=FALSE, message=FALSE, warning = FALSE}
#Compare regression models using stargazer
stargazer(model1,model2,model3,model4, type="text", digits = 2, 
          title = "Table 6: Regression Comparisons)", style = "qje",
          column.labels = c('OLS', "OLS - 3 Lags", "OLS - 2 Lags", "OLS - 1 Lag"),
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", "mortgage_rate_lag2", "mortgage_rate_lag3", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)", "Lagged Mortgage Rate (2)",
          "Lagged Mortgage Rate (3)", "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

```

The above table depicts all of the models side-by-side for comparison  Our preferred model, Model 4, is show in the column furthest right.

# Conclusion

This project examined how mortgage interest rates, housing prices, and related economic and housing market variables are associated with homeownership rates across U.S. states over time. To answer this question, we constructed a state-year panel dataset combining homeownership rates, median income, population, rent burden, mortgage rates, house price index values, unemployment rates, and housing permits per person. We used descriptive visualizations, correlation analysis, state-level distribution plots, multicollinearity checks, and regression models to study both broad trends and more specific relationships among these variables.

The analysis shows that homeownership rates vary meaningfully across both time and states. Average homeownership declined from the mid-2000s into the mid-2010s before partially recovering in the late 2010s and early 2020s. However, this recovery occurred alongside a sharp rise in house price index values and, after 2021, a rapid increase in mortgage rates. This suggests that the current housing environment is shaped by multiple affordability pressures at once: households face not only higher home prices, but also higher borrowing costs.

Several patterns from the visual analysis help clarify the broader story. The correlation heatmap and scatterplots suggest that homeownership tends to be lower in states with higher population, higher rent burden, higher house price index values, and higher unemployment. Rent burden appears especially important because it connects the rental and ownership sides of the housing market: states where renters are already financially strained may also be states where households have more difficulty moving into ownership. The box plot by state also shows substantial geographic variation, with states such as New York, California, Nevada, Hawaii, and Rhode Island consistently showing lower homeownership rates than many other states.

The regression analysis adds further support to the idea that homeownership is influenced by several overlapping economic and housing market conditions. Mortgage rates are generally negatively associated with homeownership, especially in the baseline and controlled models, which is consistent with the idea that higher borrowing costs reduce access to ownership. Housing supply, measured through housing permits per person, appears positively associated with homeownership in the models, suggesting that states with more residential construction activity may have stronger ownership outcomes. Rent burden and unemployment also provide important context, although some relationships are more complex once controls and lagged variables are included.

There are several limitations to this analysis. First, the project uses observational data, so the results should be interpreted as associations rather than definitive causal effects. Second, some variables are measured at the national level, such as the 30-year mortgage rate, while others vary by state and year, which limits how precisely we can isolate state-specific mortgage effects. Third, the exclusion of 2020 due to ACS comparability issues helps maintain consistency but removes an important year from the housing market timeline. Finally, our housing supply measure captures permits rather than completed housing units, so it may not fully reflect the actual availability of homes.

Future work could improve the analysis by incorporating more localized data at the county or metropolitan level, adding measures of housing inventory and new construction completions, and exploring differences by household income, race, age, or first-time buyer status. Additional modeling could also test fixed effects more deeply, examine regional differences, or use causal inference methods to better isolate the effects of policy-relevant housing market changes.

# Policy Recommendation
Key trends over the past two decades present an important policy concern: even though homeownership partially recovered after the mid-2010s, households now face a combination of elevated housing prices and higher borrowing costs. Given the robust finding that more permits per capita are associated with higher rates of homeownership and high rent burden negative correlation with the homeownership rate, we would advice HUD Secretary Scott Turner to approach subsequent policy with a two-pronged approach: federal legislation efforts should first be directed toward reducing zoning regulations to increase the housing supply and, subsequently, be put toward the HOME Program to allow for both financing of the building of affordable home and rental properties as well as increased access to subsidized and/or lower fixed-rate mortgages. 

To accomplish this, policy should focus on the reduction of various barriers. For example, working to update zoning regulations to streamline the use of land and permit more density and innovative construction techniques/housing types. It is important not simply to increase housing generally but to increase affordable housing, as homeownership rates are lower in largely populated areas where the rent burden is already high. In addition to increasing the homeownership rate, increasing access to affordable housing also has a wealth of other benefits for people. There have been several initiatives at the state level that can be applied at the federal level to improve homeownership generally, as well as to help states that are at a lower rate potentially catch up. Actions like lessening parking requirements, density limits, and minimum lot sizes for new residential buildings allow for more financially feasible constructions. Thus, finding ways to reduce these kinds of restrictive zoning, which are otherwise contributing to the affordability crisis in the US, may further aid this endeavor to increase homeownership rates.

By reducing zoning barriers, an important, short-term effect is being able to issue more housing permits, which can, in turn, be used to build at a lower cost per-unit. While generally increasing the supply of affordable housing supply may also contribute to a reduction in experienced rent burden, other reductions, such as in parking requirements, could contribute to lower rent and, in turn, rent burden. Doing so allows a city to meet the actual needs of a given community while lowering the costs of new construction–a burden that would otherwise be put on future tenants. In the long term, people may become more able to save for and purchase houses. Further, with more high-density housing and less parking–particularly if this is focused in transit hubs–people may be more inclined and able to opt to use public transportation and/or live closer to their jobs, which could help reduce carbon emissions in the long term. While changes to zoning to allow for densification may increase property value, however, care must be taken to ensure that low-income neighborhoods affected by these changes don’t experience high levels of gentrification and displacement–a potential, negative, and long-term effect of this process; given the homeownership rate’s relationship with rent burden, providing affordable housing must stay at the core of these policy changes. As housing supply eventually increases, high-density areas may also experience greater strain on institutions and community organizations, such as schools, hospitals, playgrounds, and places of worship.  

Furthermore, with high rent burden correlated with low homeownership, actions should be taken to also address this renter issue to allow for mobility toward homeownership for those who want to purchase a house but have historically financially struggled to do so because of the already large burden of rent. This could be approached by the development of more affordable dedicated-rental properties and/or expanded rental assistance. Given this, we also need to turn to the impact mortgage rates have, with a one percentage point increase resulting in both an immediate and long-term negative effect on homeownership. With mortgage rates rising and staying high in recent years, it would also be beneficial to expand access to and opportunity for subsidized and/or lower fixed-rate mortgages. Combined, further resources and energy need to be directed toward the HOME Investment Partnerships Program, both at the rental (renter and rental construction development) and homeownership levels. It is recommended that specific focus be given to states with the lowest rates of homeownership, like California, New York, Nevada, and Hawaii. Those at the federal level should aim to work with and hear from homeowners, developers, and local and state legislators to best understand the needs of those in areas struggling most, as well as to assist with increasing the supply of housing by supporting the implementation of pre-reviewed housing designs and planning efforts. Inspiration can also be taken from successful state policies, especially in states with higher rates of homeownership, more permits per capita, and lower rent burden.

Increased access to subsidized and/or lower fixed-rate mortgages may make homeownership more attainable for people who previously found it out of reach. By focusing the HOME Program support first to states with the lowest homeownership rates, more resources can be put toward the development of necessary intermediate steps like lowering rent burden and increasing housing supply. It may also be mutually beneficial to engage in conversation with those who have a stake in the outcome of this legislation. Community members and legislators can help shape certain decisions by providing their specific insights, and proactive involvement may help reduce the level of uncertainty or resistance residents might otherwise feel if outsiders were making decisions for and about their community. In the long-term, it is essential that the housing supply is able to increase as these kinds of mortgages become available to avoid the inflation risk it would otherwise run. An anti-displacement strategy is also required in the long run to protect current renters and homeowners in low-income communities, who may otherwise be harmed by the effects of new construction and a high density of new units. 

Ultimately, legislation that is two-fold–i.e., that addresses the issues of housing supply and affordability by reducing zoning regulations and directing HOME Program resources toward states with lower homeownership for permits and construction as well as toward increased access to subsidized and/or lower fixed-rate mortgages–will allow the Department of Housing and Urban Development to comprehensively address these barriers to homeownership.


# Recorded Policy Brief

TO BE ADDED BY MAY 11.

# Version Control

[Version control text here.]
### Team Workflow
To keep the repository organized, team members should:
1. Pull the most recent version of the repository before starting work
2. Make changes locally in the appropriate files
3. Knit `README.Rmd` after updating it so that `README.md` stays current
4. Commit changes with clear commit messages
5. Push updates after meaningful progress


## References

### Data

Federal Housing Finance Agency. (n.d.). *FHFA House Price Index® datasets*. Retrieved 2026, from https://www.fhfa.gov/data/house-price-index

Federal Reserve Bank of St. Louis. (n.d.). *30-year fixed rate mortgage average in the United States [MORTGAGE30US]*. FRED. Retrieved 2026, from https://fred.stlouisfed.org/series/MORTGAGE30US

U.S. Bureau of Labor Statistics. (n.d.). *Local Area Unemployment Statistics*. Retrieved 2026, from https://www.bls.gov/lau/

U.S. Census Bureau. (n.d.). *American Community Survey data via API*. Retrieved 2026, from https://www.census.gov/programs-surveys/acs/data/data-via-api.html

U.S. Census Bureau. (n.d.). *Building Permits Survey*. Retrieved 2026, from https://www.census.gov/permits

U.S. Department of Housing and Urban Development. (2026, May 2). HOME Investment Partnerships Program. hud.gov

### Policy

Developing an anti-displacement strategy - Local Housing Solutions. Housing Solutions Lab. (n.d.). https://www.localhousingsolutions.org/plan/developing-an-anti-displacement-strategy/ 

Freemark, Y. (2026, March 3). How zoning fits into a national housing affordability strategy | Urban Institute. Urban Institute. https://www.urban.org/urban-wire/how-zoning-fits-national-housing-affordability-strategy 

Hanley, A. (2023, December 22). Rethinking Zoning to Increase Affordable Housing. The National Association of Housing and Redevelopment Officials (NAHRO). https://www.nahro.org/journal_article/rethinking-zoning-to-increase-affordable-housing/ 

HOME Homeownership. HUD Exchange. (n.d.). https://www.hudexchange.info/programs/home/topics/homeownership/#policy-guidance-and-faqs  

HOME Rental Housing. HUD Exchange. (n.d.). https://www.hudexchange.info/programs/home/topics/rental-housing/#policy-guidance-and-faqs 

Housing reform in the states: A menu of options for 2026. Mercatus Center. (2025, September 9). https://www.mercatus.org/research/policy-briefs/housing-reform-states-menu-options-2026

Hoyt, H., & Schuetz, J. (2020, June 2). Parking requirements and foundations are driving up the cost of multifamily housing | joint center for housing studies. Harvard Joint Center for Housing Studies. https://www.jchs.harvard.edu/blog/parking-requirements-and-foundations-are-driving-up-the-cost-of-multifamily-housing 

Rifkin, C. (2024, June 7). Increasing the housing supply by reducing costs and barriers. National Conference of State Legislatures. https://www.ncsl.org/human-services/increasing-the-housing-supply-by-reducing-costs-and-barriers 
