---
title: "Class Project"
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document: default
---


### **This landing page displays the knitted output of our README.Rmd file. For the code behind the analysis and figures shown below, please consult the README.Rmd file.**

# Repository Guide

This repository contains our group project for Software Tools for Data Analysis. Our project studies how mortgage interest rates and housing prices affect homeownership rates across U.S. states over time. The final audience for this work is Scott Turner, U.S. Secretary of Housing and Urban Development (HUD).

### Branch Structure
- `main`: current working branch for the final project deliverables
- `Checkpoint-1` branch: preserved as a historical archive of our first checkpoint submission
- `Checkpoint-2` branch: preserved as a historical archive of our second checkpoint submission

### Repository Contents
- `README.Rmd`: source file for this repository landing page
- `README.md`: knitted GitHub-facing version of the README
- `Data/`: raw and cleaned data files used in the analysis
- `Report Outline.docx`: working outline for the written report
- `README_files/`: figures generated from the README

### Team Workflow
To keep the repository organized, team members should:
1. Pull the most recent version of the repository before starting work
2. Make changes locally in the appropriate files
3. Knit `README.Rmd` after updating it so that `README.md` stays current
4. Commit changes with clear commit messages
5. Push updates after meaningful progress

### Current Project Responsibilities
- Ryan: Introduction, Data Summary, README/repository instructions
- Levi: Data Analytics
- Liz: Conclusion, Policy Recommendations, repository/class page coordination

This README serves as the main landing page for our project progress, documentation, meeting notes, and selected analysis outputs.


```{r clear_environment, include=FALSE}
remove(list=ls())
```

# Introduction

Homeownership remains one of the most important pathways to household wealth-building and long-term financial stability in the United States. At the same time, recent increases in mortgage interest rates and housing prices have made homeownership less attainable for many households, raising important questions about affordability and access across states. In this report, we examine how changes in mortgage rates and housing prices relate to homeownership rates over time, with a particular focus on variation across U.S. states. By identifying the economic factors most strongly associated with homeownership outcomes, we aim to provide evidence that can help inform housing policy decisions at the federal level.

Our analysis is designed to support the policy goals of the U.S. Department of Housing and Urban Development by clarifying the extent to which borrowing costs, housing market conditions, and broader economic variables may influence homeownership. We are especially interested in whether rising mortgage rates and rising housing prices place measurable downward pressure on homeownership rates, even after accounting for other relevant state-level conditions. The findings of this report will ultimately be used to develop practical policy recommendations that can help improve housing affordability and expand access to homeownership for American households.


# Data Summary

To study these relationships, we construct a state-year panel dataset covering U.S. states from 2005 to 2024, excluding 2020 because one-year American Community Survey data were not available in a comparable form during the COVID-19 period. Each observation in the dataset represents one state in one year. This structure allows us to compare both differences across states and changes within states over time, making it well suited for analyzing how economic and housing market conditions are associated with homeownership outcomes.

The dataset combines several reputable public sources. Homeownership rates, median income, population, and rent burden are drawn from the U.S. Census Bureau’s American Community Survey. Mortgage rates come from the Federal Reserve Economic Data database, while state-level housing price data come from the Federal Housing Finance Agency House Price Index. We also include unemployment data from the Bureau of Labor Statistics and housing permit data from the Census Building Permits Survey as a proxy for housing supply. After importing, cleaning, and merging these sources, we use the resulting dataset to examine patterns in homeownership, affordability, labor market conditions, and housing supply across states. This combined dataset provides the foundation for both our descriptive analysis and our regression modeling.


```{r load_packages, message=FALSE, warning=FALSE, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))
# Load required packages
packages <- c(
  # data manipulation
  "tidyverse",
  "janitor",
  # data import/extraction
  "readr",
  "readxl",
  "car",
  "tidycensus",
  "fredr",
  # dates and time handling
  "lubridate",
  # time series structure and analysis
  "tsibble",
  "fpp3",
  "patchwork",
  # regression
  "sandwich",
  "lmtest",
  "stargazer",
  "modelsummary"
)

# checks if package available (if not - install packages)
# load packages
check.packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}
check.packages(packages)
```


```{r set_api_keys, include=FALSE}
### **Set API Keys:**
#tidycensus api key: https://api.census.gov/data/key_signup.html
#fredr api key: https://fred.stlouisfed.org/docs/api/api_key.html

census_api_key("894b3464183cec8f6c0fa053eb34d0d878de682e")
fredr_set_key("8351bf4f459d0e8c92d56f50021947d8")
```


```{r import_data, message=FALSE, warning=FALSE, include=FALSE}
#Set years to include the last two decades, excluding 2020 due to the COVID-19 pandemic.
years <- c(2005:2019, 2021:2024)

#Extract home ownership data from the Census
homeownership <- map_dfr(
  years,
  function(yr) {
    get_acs(
      geography = "state",
      variables = c(
        total_occupied = "B25003_001",
        owner_occupied = "B25003_002"
      ),
      year = yr,
      survey = "acs1"
    ) %>%
      select(NAME, variable, estimate) %>%
      pivot_wider(names_from = variable, values_from = estimate) %>%
      mutate(
        year = yr,
        homeownership_rate = 100 * owner_occupied / total_occupied
      ) %>%
      transmute(
        state = NAME,
        year,
        homeownership_rate
      )
  }
)

#Extract income data from the Census
income_list <- map(
  years,
  ~ get_acs(
      geography = "state",
      variables = "B19013_001",
      year = .x,
      survey = "acs1"
    ) %>%
      transmute(
        state = NAME,
        year = .x,
        median_income = estimate
      )
)

#Bind the income data into a DF
income <- bind_rows(income_list)

#Extract population data from the Census
population_list <- map(
  years,
  ~ get_acs(
      geography = "state",
      variables = "B01003_001",
      year = .x,
      survey = "acs1"
    ) %>%
      transmute(
        state = NAME,
        year = .x,
        population = estimate
      )
)

#Bind the population data into a DF
population <- bind_rows(population_list)

# Rent burden from ACS table B25070
# Defined as share of renter households paying 30% or more of income toward gross rent
rent_burden <- map_dfr(
  years,
  function(yr) {
    get_acs(
      geography = "state",
      table = "B25070",
      year = yr,
      survey = "acs1"
    ) %>%
      select(NAME, variable, estimate) %>%
      mutate(year = yr) %>%
      pivot_wider(names_from = variable, values_from = estimate) %>%
      transmute(
        state = NAME,
        year,
        rent_burden = 100 * (
          B25070_007 + B25070_008 + B25070_009 + B25070_010
        ) / B25070_001
      )
  }
)

# Housing supply from Census Building Permits Survey
# Read directly from Census-hosted Excel file
permits_url <- "https://www.census.gov/construction/bps/xls/annualhistorybystate.xls"

tmp_permits <- tempfile(fileext = ".xls")
download.file(permits_url, tmp_permits, mode = "wb")

sheet_names <- excel_sheets(tmp_permits)
sheet_names <- sheet_names[sheet_names != "PR"]

permits_state <- map_dfr(
  sheet_names,
  function(sh) {
    read_excel(
      tmp_permits,
      sheet = sh,
      skip = 5,
      col_names = FALSE
    ) %>%
      select(1:5) %>%
      setNames(c(
        "universe",
        "year",
        "total_buildings",
        "housing_permits",
        "total_valuation"
      )) %>%
      filter(
        !is.na(year),
        str_detect(as.character(year), "^[0-9]{4}$")
      ) %>%
      transmute(
        state = state.name[match(sh, state.abb)],
        year = as.integer(year),
        housing_permits = readr::parse_number(as.character(housing_permits))
      ) %>%
      filter(year %in% years)
  }
)

#Extract mortgage data from FRED
mortgage_raw <- fredr(
  series_id = "MORTGAGE30US",
  observation_start = as.Date("2005-01-01"),
  observation_end = Sys.Date()
)

#Estimate annual mortgage rates
mortgage_annual <- mortgage_raw %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarise(
    mortgage_rate = mean(value, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r import_local_data, message=FALSE, warning=FALSE, include=FALSE}
# Import Local Data
fhfa_raw <- read_csv("Data/hpi_at_state (1).csv", col_names = FALSE) %>%
  setNames(c("state", "year", "quarter", "hpi"))

# Set HPI data by state/year
fhfa_annual <- fhfa_raw %>%
  clean_names() %>%
  filter(year >= 2005) %>%
  group_by(state, year) %>%
  summarise(
    hpi = mean(hpi, na.rm = TRUE),
    .groups = "drop"
  )

hpi_annual <- fhfa_annual %>%
  mutate(state = state.name[match(state, state.abb)])

# Import local unemployment data
bls_data <- read_delim("Data/la.data.2.AllStatesU.txt", delim = "\t") %>%
  clean_names()

bls_series <- read_delim("Data/la.series.txt", delim = "\t") %>%
  clean_names()

bls_area <- read_delim("Data/la.area", delim = "\t") %>%
  clean_names()

# Join the actual data to the series
bls_merged <- bls_data %>%
  left_join(bls_series, by = "series_id")

# Use key to select code
bls_merged <- bls_merged %>%
  left_join(
    bls_area %>% select(area_code, area_text),
    by = "area_code"
  )

# Keep annual-average unemployment rate
state_unemployment <- bls_merged %>%
  filter(period == "M13") %>%          # annual average
  filter(measure_code == "03") %>%     # unemployment rate
  transmute(
    state = area_text,
    year = as.integer(year),
    unemployment_rate = as.numeric(value)
  ) %>%
  filter(state %in% c(state.name, "District of Columbia")) %>%
  arrange(state, year)
```



```{r merge_data, message=FALSE, warning=FALSE, include=FALSE}
#Merge the data sets by year and state
merged <- left_join(
  population,
  income,
  by = c("state", "year")
) %>%
  left_join(
    homeownership,
    by = c("state", "year")
  ) %>%
  left_join(
    hpi_annual,
    by = c("state", "year")
  ) %>%
  left_join(
    state_unemployment,
    by = c("state", "year")
  ) %>%
  left_join(
    rent_burden,
    by = c("state", "year")
  ) %>%
  left_join(
    permits_state,
    by = c("state", "year")
  ) %>%
  left_join(
    mortgage_annual,
    by = "year"
  )
```


```{r permits_per_person, message=FALSE, warning=FALSE, include=FALSE}
#Estimate annual mortgage rates
merged <- merged %>%
  mutate(perm_per_person = housing_permits/population, NA)
```


```{r excludeDCandPR, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Filter out D.C. and PR
merged <- merged %>%
  filter(!state %in% c("Puerto Rico", "District of Columbia"))

#Remove a bad column
merged <- merged[, names(merged) != "NA"]
```


```{r pop_summary_stats, echo=FALSE, warning=FALSE, message=FALSE}
#Make a df version of merged
merged_df <- as.data.frame(merged)

#Rename quantitative variables in merged_df
merged_df <- merged_df %>%
       select('Population' = population,
              "Median Income" = median_income,
              "Homeownership Rate" = homeownership_rate,
              "Mortgage Rate" = mortgage_rate,
              "Housing Price Index" = hpi,
              "Unemployment Rate" = unemployment_rate,
              "Rent Burden" = rent_burden,
              "Housing Permits" = housing_permits,
              "Housing Permits Per Person" = perm_per_person)

#Compute summary statistics for all quantitative variables
datasummary_skim(merged_df[, sapply(merged_df, is.numeric)], histogram=FALSE, title = "Summary Statistics", align="c")

```


[Discussion of summary stats.]


### **Plot Histograms of Key Variables:**
```{r figure_1, echo=FALSE, message=FALSE, warning = FALSE}
key_vars_labs <- c("homeownership_rate" = "Homeownership Rate", "hpi" = "House Price Index", "median_income" = "Median Income", "population" = "Population", "rent_burden" = "Rent Burden", "unemployment_rate" = "Unemployment Rate", "perm_per_person" = "Housing Permits per Person")

#Create histogram plots of key variables
merged %>%
  select(
    homeownership_rate,
    median_income,
    population,
    hpi,
    unemployment_rate,
    rent_burden,
    perm_per_person
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 2, 
             labeller = as_labeller(key_vars_labs)) +
  labs(
    title = "Histograms of Key Variables",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal()
```

[Discussion of histograms.]

# Data Analytics

### **Key Trends Over Time:**
```{r figure_2, echo=FALSE, message=FALSE, }
# Average homeownership rate over time
avg_homeownership_ts <- merged %>%
  group_by(year) %>%
  summarise(
    avg_homeownership_rate = mean(homeownership_rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as_tsibble(index = year)

# Plot using fpp3
p1 <- autoplot(avg_homeownership_ts, avg_homeownership_rate) +
  labs(
    title = "Average State Homeownership Rate Over Time",
    x = "Year",
    y = "homeownership Rate (%)"
  )

# Average HPI over time
avg_hpi_ts <- merged %>%
  group_by(year) %>%
  summarise(
    avg_hpi = mean(hpi, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as_tsibble(index = year)

# Plot using fpp3
p2 <- autoplot(avg_hpi_ts, avg_hpi) +
  labs(
    title = "Average State House Price Index Over Time",
    x = "Year",
    y = "Average HPI"
  )

# Mortgage rate over time
mortgage_ts <- mortgage_annual %>%
  as_tsibble(index = year)

# Plot using fpp3
p3 <- autoplot(mortgage_ts, mortgage_rate) +
  labs(
    title = "Average 30-Year Mortgage Rate Over Time",
    x = "Year",
    y = "Mortgage Rate (%)"
  )

# Combine all three plots
p1 / p2 / p3

```

[Discussion of trends over time.]

### **Correlation Matrix:**
```{r figure_3, echo=FALSE, message=FALSE}
#Create correlation matrix
#Select variables of interest
corr_data <- merged %>%
  select(
    homeownership_rate,
    median_income,
    population,
    hpi,
    unemployment_rate,
    mortgage_rate,
    rent_burden,
    perm_per_person
  )

corr_matrix <- cor(corr_data, use = "complete.obs")

corr_df <- as.data.frame(as.table(corr_matrix))

# labels
var_labels <- c(
  homeownership_rate = "Homeownership Rate",
  median_income      = "Median Income",
  population         = "Population",
  hpi                = "House Price Index",
  unemployment_rate  = "Unemployment Rate",
  mortgage_rate      = "Mortgage Rate",
  rent_burden        = "Rent Burden",
  perm_per_person    = "Housing Permits per Person"
)

# keep only upper triangle, relabel
corr_df <- corr_df %>%
  mutate(
    Var1 = factor(Var1, levels = colnames(corr_matrix)),
    Var2 = factor(Var2, levels = colnames(corr_matrix))
  ) %>%
  filter(as.numeric(Var1) <= as.numeric(Var2)) %>%
  mutate(
    Var1 = factor(var_labels[as.character(Var1)], levels = var_labels),
    Var2 = factor(var_labels[as.character(Var2)], levels = var_labels)
  )

#Plot the correlation matrix
ggplot(corr_df, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", Freq)), size = 3.5) +
  scale_fill_gradient2(
    low = "firebrick",
    mid = "white",
    high = "steelblue",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation Heatmap of Key Variables",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

[Discussion of correlation matrix.]

```{r missing_check, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Check for missing obsevations
merged %>%
  summarise(
    across(
      c(
        homeownership_rate,
        median_income,
        population,
        hpi,
        unemployment_rate,
        mortgage_rate,
        rent_burden,
        perm_per_person
      ),
      ~ sum(is.na(.))
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  )
```


```{r figure_4, echo=FALSE, message=FALSE, }
#Create lines of best fit for key variables
#Identify key variables
kpi_labs <- c("hpi" = "House Price Index",
              "median_income" = "Median Income", 
              "mortgage_rate" = "Mortgage Rate", 
              "perm_per_person" = "Housing Permits per Person", 
              "rent_burden" = "Rent Burden", 
              "unemployment_rate" = "Unemployment Rate")

#Plot the lines of best fit
merged %>%
  select(
    homeownership_rate,
    mortgage_rate,
    hpi,
    unemployment_rate,
    median_income,
    rent_burden,
    perm_per_person
  ) %>%
  pivot_longer(
    cols = -homeownership_rate,
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value, y = homeownership_rate)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~ variable, scales = "free_x", ncol = 2, labeller = as_labeller(kpi_labs)) +
  labs(
    title = "Homeownership Rate vs. Key Predictors",
    x = "Predictor",
    y = "Homeownership Rate",
  ) +
  theme_minimal()
```

[Discussion of lines of best fit.]

```{r figure_5, echo=FALSE, message=FALSE, }
#Create box plot for homeownership rate by state
merged %>%
  ggplot(aes(x = reorder(state, homeownership_rate, median), y = homeownership_rate)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Homeownership Rate by State",
    x = "State",
    y = "Homeownership Rate"
  ) +
  theme_minimal()
```

[Discussion of box and whisker plots.]

### **Multicollinearity Check:**
```{r multicollinearity_check, echo=FALSE, message=FALSE, }
#Check for multicolinearity
vif_model <- lm(
  homeownership_rate ~ mortgage_rate + hpi + unemployment_rate +
    median_income + rent_burden + perm_per_person,
  data = merged
)

vif(vif_model)
```

[Discussion of multicollinearity check.]

```{r lag_variables, message=FALSE, warning=FALSE, include=FALSE}
#Create lag variables
merged <- merged %>%
  arrange(state, year) %>%
  group_by(state) %>%
  mutate(
    mortgage_rate_lag1 = lag(mortgage_rate, 1),
    mortgage_rate_lag2 = lag(mortgage_rate, 2),
    mortgage_rate_lag3 = lag(mortgage_rate, 3),
    hpi_lag1 = lag(hpi, 1),
    hpi_lag2 = lag(hpi, 2),
    hpi_lag3 = lag(hpi, 3)
  ) %>%
  ungroup()
```

## **Regression Analysis**
For directions on how to conduct multilinear regression and plot confidence intervals, [click here](https://www.geeksforgeeks.org/r-language/how-to-use-the-coeftest-function-in-r/).

### **Basic Model**
```{r regression_basic, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model <- lm(homeownership_rate ~ mortgage_rate + state + 
               as.factor(year), data = merged )

#Summarize model
stargazer(model, type="text", digits = 2, 
          title = "Table 1: Simple OLS Regression", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate"), 
          covariate.labels=c("Mortgage Rate"))

#Create coefficients and standard errors
coefficients <- coef(model)[c("mortgage_rate")]
std_errors <- sqrt(diag(vcov(model)))[c("mortgage_rate")]

#Create data needed to plot confidence intervals
plot_data <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error, ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels = c(mortgage_rate = "Mortgage Rate"))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

[Discussion of basic model.]

### **Model 1**
```{r regression_1, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model1 <- lm(homeownership_rate ~ mortgage_rate + hpi + median_income +
               unemployment_rate + rent_burden + perm_per_person + state + 
               as.factor(year), data = merged )

#Summarize model
stargazer(model1, type="text", digits = 2, 
          title = "Table 2: OLS Regression with Controls", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "House Price Index", "Median Income", "Unemployment Rate",
          "Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model1)[c("mortgage_rate", "hpi", "median_income", 
                               "unemployment_rate", "rent_burden")]
std_errors <- sqrt(diag(vcov(model1)))[c("mortgage_rate", "hpi", 
                                         "median_income", "unemployment_rate", 
                                         "rent_burden")]

#Create data needed to plot confidence intervals
plot_data1 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data1, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels = c(
    hpi = "House Price Index",
    median_income = "Median Income",
    mortgage_rate = "Mortgage Rate",
    rent_burden = "Rent Burden",
    unemployment_rate = "Unemployment Rate"
  ))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

```{r clustered_standarderrors, echo=FALSE, message=FALSE, }
#Calculate clustered standard errors for Model 1
model1_cluster_se <- vcovCL(model1, cluster = ~ state)
summary_clustered <- coeftest(model1, vcov = model1_cluster_se)


```

[Discussion of model 1.]

### **Model 2**
```{r regression_2, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model2 <- lm(homeownership_rate ~ mortgage_rate + mortgage_rate_lag1 + 
               mortgage_rate_lag2 + mortgage_rate_lag3 + hpi + median_income +
               unemployment_rate + rent_burden + perm_per_person + 
               state + as.factor(year), data = merged )

#Summarize model
stargazer(model2, type="text", digits = 2, 
          title = "Table 3: OLS Regression with 3 Lags", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", "mortgage_rate_lag2", "mortgage_rate_lag3", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)", "Lagged Mortgage Rate (2)",
          "Lagged Mortgage Rate (3)", "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model2)[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "mortgage_rate_lag3", "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]
std_errors <- sqrt(diag(vcov(model2)))[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "mortgage_rate_lag3", "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]

#Create data needed to plot confidence intervals
plot_data2 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data2, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels = c("hpi" = "House Price Index",
                   "median_income" = "Median Income",
                   "mortgage_rate"= "Mortgage Rate", 
                   "mortgage_rate_lag1"= "Lagged Mortgage Rate (1)",
                   "mortgage_rate_lag2"= "Lagged Mortgage Rate (2)",
                   "mortgage_rate_lag3"= "Lagged Mortgage Rate (3)",
                   "rent_burden" = "Rent Burden",
                   "unemployment_rate" = "Unemployment Rate"))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

[Discussion of model 2.]

### **Model 3**
```{r regression_3, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model3 <- lm(homeownership_rate ~ mortgage_rate + mortgage_rate_lag1 + 
               mortgage_rate_lag2 + hpi + median_income +
               unemployment_rate + rent_burden + perm_per_person + state +
               as.factor(year), data = merged )

#Summarize model
stargazer(model2, type="text", digits = 2, 
          title = "Table 4: OLS Regression with 2 Lags", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", "mortgage_rate_lag2", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)", "Lagged Mortgage Rate (2)",
          "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model3)[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]
std_errors <- sqrt(diag(vcov(model3)))[c("mortgage_rate", "mortgage_rate_lag1", "mortgage_rate_lag2", 
                               "hpi", "median_income", "unemployment_rate",
                               "rent_burden")]

#Create data needed to plot confidence intervals
plot_data3 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data3, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  theme_minimal() +
  scale_x_discrete(labels=c("hpi" = "House Price Index",
                   "median_income" = "Median Income",
                   "mortgage_rate"= "Mortgage Rate", 
                   "mortgage_rate_lag1"= "Lagged Mortgage Rate (1)",
                   "mortgage_rate_lag2"= "Lagged Mortgage Rate (2)",
                   "rent_burden" = "Rent Burden",
                   "unemployment_rate" = "Unemployment Rate"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

[Discussion of model 3.]

### **Model 4**
```{r regression_4, echo=FALSE, message=FALSE, warning = FALSE}
#Create a simple multilinear regression model
model4 <- lm(homeownership_rate ~ mortgage_rate + mortgage_rate_lag1 + 
               hpi + median_income + unemployment_rate + rent_burden + 
               perm_per_person + state + as.factor(year), data = merged )

#Summarize model
stargazer(model2, type="text", digits = 2, 
          title = "Table 5: OLS Regression with 1 Lag", style = "qje",
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)",
          "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

#Create coefficients and standard errors
coefficients <- coef(model4)[c("mortgage_rate", "mortgage_rate_lag1", "hpi", 
                               "median_income", "unemployment_rate",
                               "rent_burden")]
std_errors <- sqrt(diag(vcov(model4)))[c("mortgage_rate", "mortgage_rate_lag1", "hpi", "median_income", "unemployment_rate", "rent_burden")]

#Create data needed to plot confidence intervals
plot_data4 <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Std_Error = std_errors
)

#create plot of coefficients and confidence interval
ggplot(plot_data4, aes(x = Coefficient, y = Estimate, ymin = Estimate - 1.96 * Std_Error,
                                           ymax = Estimate + 1.96 * Std_Error)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  labs(title = "Coefficients and Confidence Intervals",
       x = "Coefficient",
       y = "Estimate") +
  scale_x_discrete(labels=c("hpi" = "House Price Index",
                   "median_income" = "Median Income",
                   "mortgage_rate"= "Mortgage Rate", 
                   "mortgage_rate_lag1"= "Lagged Mortgage Rate (1)",
                   "rent_burden" = "Rent Burden",
                   "unemployment_rate" = "Unemployment Rate"))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

[Discussion of model 4.]

### **Compare Models**
```{r regression_comparison, echo=FALSE, message=FALSE, warning = FALSE}
#Compare regression models using stargazer
stargazer(model1,model2,model3,model4, type="text", digits = 2, 
          title = "Table 6: Regression Comparisons)", style = "qje",
          column.labels = c('OLS', "OLS - 3 Lags", "OLS - 2 Lags", "OLS - 1 Lag"),
          dep.var.labels = "Homeownership Rate", keep=c("mortgage_rate", 
          "mortgage_rate_lag1", "mortgage_rate_lag2", "mortgage_rate_lag3", 
          "hpi", "median_income", "unemployment_rate","rent_burden", 
          "perm_per_person"), covariate.labels=c("Mortgage Rate", 
          "Lagged Mortgage Rate (1)", "Lagged Mortgage Rate (2)",
          "Lagged Mortgage Rate (3)", "House Price Index", "Median Income",
          "Unemployment Rate","Rent Burden", "Housing Permits per Person"))

```

[Comparison of models and selection of model 4 as preffered model.]

# Conclusion

[Conclusion text here.]

# Policy Recommendation

Turning to policy, it should first and foremost be the priority to increase the housing supply, given the robust finding that more permits per capita result in higher rates of homeownership. To do so, certain barriers can be reduced. For example, working to update zoning regulations to streamline the use of land and permit more density and innovative construction techniques/housing types. It is important not simply to increase housing generally but to increase affordable housing, as homeownership rates are lower in largely populated areas where the rent burden is already high. In addition to increasing the homeownership rate, increasing access to affordable housing also has a wealth of other benefits for people. There have been several initiatives at the state level that can be applied at the federal level to improve homeownership generally, as well as to help states that are at a lower rate potentially catch up. Actions like lessening parking requirements, density limits, and minimum lot sizes for new residential buildings allow for more financially feasible constructions. Thus, finding ways to reduce these kinds of restrictive zoning, which are otherwise contributing to the affordability crisis in the US, may further aid this endeavor to increase homeownership rates.


With high rent burden correlated with low homeownership, actions should be taken to also address this renter issue to allow for mobility toward homeownership for those who want to purchase a house but have historically financially struggled to do so because of the already large burden of rent. This could be approached by the development of more affordable rental properties and/or expanding rental assistance.  We also need to turn to the impact mortgage rates have, with a one percentage point increase resulting in both an immediate and long-term negative effect on homeownership. With mortgage rates rising and staying high in recent years, it may be helpful to expand access to and opportunity for fixed-rate mortgages at lower rates. Combined, further resources and energy need to be directed toward the HOME Investment Partnerships Program, both at the rental (renter and rental construction development) and homeownership levels.

Specific focus should be given to states with the lowest rates of homeownership, like California, New York, Nevada, and Hawaii. Those at the federal level should aim to work with and hear from homeowners, developers, and local and state legislators to best understand the needs of those in areas struggling most. Inspiration can also be taken from successful state policies, especially in states with higher rates of homeownership, more permits per person, and lower rent burden.


# Recorded Policy Brief

TO BE ADDED BY MAY 11.

# Version Control

[Version control text here.]

# References

- https://fred.stlouisfed.org/series/MORTGAGE30US 
- https://www.fhfa.gov/data/hpi/datasets?tab=quarterly-data 
- https://www.census.gov/data.html 
- https://www.bls.gov/web/laus/laumstrk.htm 
